---
title: "ATR Analysis - ITQB"
author: "Adrianne Seiden"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: darkly
    highlight: zenburn
    code_folding: hide
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding,
                     output_dir = "../adrianne-code/") })
---
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  results = "markup",
  include = TRUE
)
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))
## Installing packages ##
list_of_packages <- c(
  "plyr",
  "dplyr",
  "stringr",
  "reshape",
  "data.table",
  "nc",
  "readr",
  "tidyr",
  "pracma"
)

# install missing packages
new_packages <- list_of_packages[!(
  list_of_packages %in% installed.packages()[, "Package"]
)]
if (length(new_packages)) install.packages(new_packages)

# load packages
library(plyr)
library(dplyr)
library(stringr)
library(reshape)
library(data.table)
library(nc)
library(readr)
library(tidyr)
library(pracma)
library(ggplot2)
library(ggridges)
library(cowplot)
library(grid)
library(ggsignif)

# Set paths
dpt_folder <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/all_abs_DPT_files"
outputs_folder <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/ATR_outputs"


# Helper function to find the most recent processed integration data file
find_most_recent_processed_file <- function(data_dir) {
  pattern <- "processed_int_data_.*\\.csv$"
  files <- list.files(data_dir, pattern = pattern, full.names = TRUE)
  
  if (length(files) == 0) {
    return(NULL)
  }
  
  # Extract dates from filenames and find the most recent
  dates <- stringr::str_extract(basename(files), "\\d{4}-\\d{2}-\\d{2}")
  most_recent_idx <- which.max(as.Date(dates))
  
  cat("Found", length(files), "processed integration files\n")
  cat("Using most recent file:", basename(files[most_recent_idx]), "\n")
  
  return(files[most_recent_idx])
}

# Source centralized theme and color configuration
source(file.path(dirname(knitr::current_input()), "config/ITQB_color_config.R"))

```

# Overview
placeholder

# Preparing data for analysis

## Importing DPT data
This code is modified from code by Stephany Soledad Chacon, "Box> Salk Institute Project> Salk Data> FTIR_Intact_decomposition_pots> FTIR_analysis.Rmd". Using data in the DPT format, we can extrapolate sample attributes from the file names, then normalize and visualize spectra across replicates, crops, and treatments. The width of the line bounding the spectra represents the range of the data.
```{r import-dpt-data}
# Check if folder exists and list files
if (!dir.exists(dpt_folder)) {
  stop("Directory does not exist: ", dpt_folder)
}

# List and sort files
cat("Files found in directory:\n")
print(list.files(dpt_folder))
DPTfiles <- sort(list.files(dpt_folder, pattern = ".dpt"))
cat("Number of .dpt files:", length(DPTfiles), "\n")

# Read files individually and combine
dpt_data_list <- list()

for (i in seq_along(DPTfiles)) {
  file_path <- file.path(dpt_folder, DPTfiles[i])
  
  # Read as lines and parse manually (DPT files are space-delimited)
  lines <- readLines(file_path, warn = FALSE)
  lines <- lines[lines != ""]  # Remove empty lines
  
  # Split each line by whitespace
  split_lines <- strsplit(lines, "\\s+")
  
  # Convert to data frame
  temp_data <- tryCatch({
    data.frame(
      wavenumber = as.numeric(sapply(split_lines, `[`, 1)),
      absorbance = as.numeric(sapply(split_lines, `[`, 2)),
      stringsAsFactors = FALSE
    )
  }, error = function(e) NULL)
  
  if (!is.null(temp_data)) {
    # Add filename column
    temp_data$filename <- DPTfiles[i]
    
    # Store in list
    dpt_data_list[[i]] <- temp_data
  }
}

# Remove NULL entries (failed reads)
dpt_data_list <- dpt_data_list[!sapply(dpt_data_list, is.null)]

# Combine all data
dpt_data <- bind_rows(dpt_data_list)

# Remove any rows with NA values
dpt_data <- dpt_data[complete.cases(dpt_data), ]

# Extract metadata from filenames

# Helper to extract fields from filenames with the pattern
# AKS_suberin_[plant]_[washing-treatment]_[yymmdd]-[run].dpt
extract_filename_fields <- function(filename) {
  stem <- tools::file_path_sans_ext(basename(filename))
  matches <- stringr::str_match(
    stem,
    "^AKS_suberin_([^_]+)_([^_]+)_([0-9]{6})-([0-9]+)$"
  )

  if (is.na(matches[1, 1])) {
    return(tibble(
      plant = NA_character_,
      method = NA_character_,
      run_date = as.Date(NA),
      run_rep = NA_integer_
    ))
  }

  tibble(
    plant = tolower(matches[1, 2]),
    method = tolower(matches[1, 3]),
    run_date = as.Date(matches[1, 4], format = "%y%m%d"),
    run_rep = as.integer(matches[1, 5])
  )
}

# Check for problems in filename parsing
bad_files <- dpt_data$filename[
  !stringr::str_detect(
    basename(dpt_data$filename),
    "^AKS_suberin_[^_]+_[^_]+_[0-9]{6}-[0-9]+\\.dpt$"
  )
]

bad_files

file_metadata <- bind_rows(lapply(dpt_data$filename, extract_filename_fields))
dpt_data <- bind_cols(dpt_data, file_metadata)

# save as csv for record-keeping
write.csv(dpt_data, file.path(outputs_folder, "dpt_data.csv"), row.names = FALSE)


cat("Samples by plant:\n")
print(table( dpt_data$plant))
cat("Samples by treatment:\n") 
print(table( dpt_data$method))

```

### Preprocessing
Define color palettes for each plant and method combination. This will be used in the plots to maintain consistent coloring across different visualizations.
```{r preprocessing}
# define color palettes for each plant and method combination
methods <- sort(unique( dpt_data$method))
plants <- sort(unique( dpt_data$plant))

plant_palette <- get_plant_palette(
  methods = methods,
  plants = plants
)

dpt_data$color_key <-
    paste(dpt_data$plant,
          dpt_data$method,
          sep = "_")

plant_colors <- get_plant_colors(sort(unique(dpt_data$plant)))

```

## Normalizing data
This code is modified to use baseline correction (subtracting the minimum) and standard normal variate (SNV) normalization.
```{r normalize-data}
# Normalize absorbance data by sample (same approach as OPUSprocessing.Rmd)
dpt_data <- dpt_data %>%
  group_by(filename) %>%
  mutate(
    # Baseline correction - subtract minimum
    baseline_corrected = absorbance - min(absorbance, na.rm = TRUE),
    # Apply SNV (Standard Normal Variate) normalization
    mean_abs = mean(baseline_corrected, na.rm = TRUE),
    sd_abs = sd(baseline_corrected, na.rm = TRUE),
    normalized_absorbance = (baseline_corrected - mean_abs) / sd_abs
  ) %>%
  select(-mean_abs, -sd_abs) %>%
  ungroup()
dpt_data <- dpt_data %>%
  group_by(filename) %>%
  mutate(
    # Baseline correction - subtract minimum
    snv_corrected = normalized_absorbance - min(normalized_absorbance, na.rm = TRUE)
  ) %>%
  ungroup()

## Save normalized data for record-keeping

write.csv(dpt_data, file.path(outputs_folder, "dpt_data_normalized.csv"), row.names = FALSE)
```


# Plotting ATR spectra
## Plotting by preparation method
These plots are of normalized absorbance values. The thickness of the opaque line represents the range of the data. Annotations indicate integration regions used for calculations of simple plant, complex plant, and microbial organic matter proportions.
Integration regions according to Mainka et al. (2022):

Mainka, M., Summerauer, L., Wasner, D., Garland, G., Griepentrog, M., Berhe, A. A., & Doetterl, S. (2022). Soil geochemistry as a driver of soil organic matter composition: Insights from a soil chronosequence. Biogeosciences, 19(6), 1675–1689. https://doi.org/10.5194/bg-19-1675-2022

Plus carbonyl region (1780 - 1690 cm<sup>-1</sup>) \& sugars region (1200 - 900 cm<sup>-1</sup>) added to the analysis.

<table style="width: auto; border-collapse: collapse; border: 0.1px solid #f5f5f5;">
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Aliphatics (simple plant):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">2976 - 2898 cm<sup>-1</sup> + 2870 - 2839 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Carbonyls:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1780 - 1690 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Microbial:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1660 - 1580 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Aromatics (complex plant):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1550 - 1500 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Sugars:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1200 - 900 cm<sup>-1</sup></td>
</tr>
</table>

```{r prep-method-plots, results='asis'}

# Generic function to create ridge plots for any preparation method
create_method_ridge_plot <- function(data, target_method, title = NULL) {
  # Filter data for specific method
  filtered_data <- data %>%
    filter(method == target_method)

  # Create title if not provided
  if (is.null(title)) {
    title <- paste("Roots (", target_method, ")", sep = "")
  }
  
  # Determine y-position for annotations based on data
  present_levels <- sort(unique(filtered_data$plant))  # only levels actually in this week's data, in original order

  max_y <- 7 - length(present_levels)  
  
  # Create the plot
  ridge_plot <- filtered_data %>%
    ggplot(mapping = aes(x = wavenumber,
                         y = plant,
                         height = snv_corrected,
                         group = plant,
                         fill = plant,
                         color = plant)) +
    geom_density_ridges(stat = "identity",
                        scale = 3, # overlap scale - 1 is no overlap, >1 is more overlap
                        alpha = 0.25) +
                        
    scale_x_reverse(
      limits = c(3750, 650),
      breaks = seq(4000, 600, -500),  # Ticks every 500 from 4000 down to 600
      expand = c(0.001, 0.001) 
      # Remove padding on x-axis
    ) +
    scale_y_discrete(expand = expansion(mult = c(0, 0.1), add = c(0, 0))) +  # space below, space above+  # space below, space above
    
    theme(axis.text.y = element_text(size = 12),
          legend.position = "none") +
    scale_fill_manual(values = plant_colors) +
    scale_color_manual(values = plant_colors) +
    ggtitle(title) +
    labs(x = expression("wavenumber (cm"^-1*")"), y = "") +
    # Add simple plant region annotation
    annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("text", x = 2900, y = max_y * 1.08, label = "Simple Plant", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add carbonyl region annotation
    annotate("rect", xmin = 1780, xmax = 1690, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#83a0e8") +
    annotate("text", x = 1910, y = max_y * 0.9, label = "Carbonyl", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add microbial region annotation
    annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#a8cbad") +
    annotate("text", x = 1640, y = max_y * 1.05, label = "Microbial", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add complex plant region annotation
    annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#beb5d5") +
    annotate("text", x = 1300, y = max_y * 1.22, label = "Complex Plant", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add sugars region annotation
    annotate("rect", xmin = 1200, xmax = 900, ymin = -Inf, ymax = Inf, 
             alpha = 0.17, fill = "#c9aac4") +
    annotate("text", x = 933, y = max_y * 1.4, label = "Sugars", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) 
  
  return(list(plot = ridge_plot, data = filtered_data))
}

# Generic function to create sample count tables
create_sample_count_table <- function(filtered_data, method) {
  col_name <- paste(method, "samples")
  
  counts <- filtered_data %>%
    select(filename, plant) %>%
    distinct() %>%
    count(plant) %>%
    mutate(!!col_name := paste(n, "&nbsp;&nbsp;", plant)) %>%
    select(all_of(col_name))
  
  return(counts)
}

# Create plots and tables for each method
methods <- sort(unique(dpt_data$method))

for (mt in methods) {
  # Create plot and get filtered data
  result <- create_method_ridge_plot(dpt_data, mt)
  
  # Print the plot
  print(result$plot)
  
  # Create and print sample count table with results='asis'
  counts_table <- create_sample_count_table(result$data, mt)
  cat(knitr::kable(counts_table, format = "html", escape = FALSE))
  cat("<br><br>")  # Add HTML line breaks for spacing
}

# troubleshooting method colors
method_levels <- sort(unique(dpt_data$method))

if (!all(method_levels %in% names(method_colors))) {
  missing_colors <- setdiff(method_levels, names(method_colors))
  cat(missing_colors," do not have defined colors in method_colors.")
}

```

## Plotting by plant
These plots are of normalized absorbance values. The thickness of the opaque line represents the range of the data. Annotations indicate integration regions used for calculations of simple plant, complex plant, and microbial organic matter proportions.  

<table style="width: auto; border-collapse: collapse; border: 0.1px solid #f5f5f5;">
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Aliphatics (simple plant):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">2976 - 2898 cm<sup>-1</sup> + 2870 - 2839 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Carbonyls:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1780 - 1690 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Microbial:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1660 - 1580 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Aromatics (complex plant):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1550 - 1500 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Sugars:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1200 - 900 cm<sup>-1</sup></td>
</tr>
</table>

```{r facet-plant-plots, results='asis'}
# check which methods do not have defined colors

# Generic function to create ridge plots for any preparation method
create_plant_faceted_plot <- function(data, current_plant, title = NULL) {
  # Filter data for specific method
  filtered_data <- data %>%
    filter(plant == current_plant)

  # Create title if not provided
  if (is.null(title)) {
    title <- paste(str_to_title(current_plant), "roots", sep = " ")
  }
  
  # Determine y-position for annotations based on data
  present_levels <- sort(unique(filtered_data$method))  # only levels actually in this week's data, in original order

  max_y <- 7 - length(present_levels)/5
  
  # Fixed method colors (from ITQB_color_config.R) -- method is now the hue,
  # not a shade of this plant's hue, so a given method is always the same
  # color here no matter which plant it's shown for.
  method_colors_current <- get_method_colors(present_levels)

  # Create the plot
  ridge_plot <- filtered_data %>%
    ggplot(mapping = aes(x = wavenumber,
                         y = method,
                         height = snv_corrected,
                         group = method,
                         fill = method,
                         color = method)) +
    geom_density_ridges(stat = "identity",
                        scale = 3, # overlap scale - 1 is no overlap, >1 is more overlap
                        alpha = 0.25) +
   scale_x_reverse(
      limits = c(3750, 650),
      breaks = seq(4000, 600, -500),  # Ticks every 500 from 4000 down to 600
      expand = c(0.001, 0.001) 
      # Remove padding on x-axis
    ) +
    scale_y_discrete(expand = expansion(mult = c(0, 0.1), add = c(0, 0))) +  # space below, space above+  # space below, space above
    
    theme(axis.text.y = element_text(size = 12),
          legend.position = "none") +
    scale_fill_manual(values = method_colors_current) +
    scale_color_manual(values = method_colors_current) +
    ggtitle(title) +
    labs(x = expression("wavenumber (cm"^-1*")"), y = "") +
    # Add simple plant region annotation
    annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("text", x = 2900, y = max_y*1.03, label = "Simple Plant", 
             size = 3.33, hjust = 0.5, color = annotation_text_color) +
    # Add carbonyl region annotation
    annotate("rect", xmin = 1780, xmax = 1690, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#83a0e8") +
    annotate("text", x = 1900, y = max_y*0.95, label = "Carbonyl", 
             size = 3.33, hjust = 0.5, color = annotation_text_color) +
    # Add microbial region annotation
    annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#a8cbad") +
    annotate("text", x = 1640, y = max_y * 1.07, label = "Microbial", 
             size = 3.33, hjust = 0.5, color = annotation_text_color) +
    # Add complex plant region annotation
    annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#beb5d5") +
    annotate("text", x = 1333, y = max_y * 1.22, label = "Complex Plant", 
             size = 3.33, hjust = 0.5, color = annotation_text_color) +
      # Add sugars region annotation
    annotate("rect", xmin = 1200, xmax = 900, ymin = -Inf, ymax = Inf, 
             alpha = 0.16, fill = "#c3aac1") +
    annotate("text", x = 933, y = max_y * 1.29, label = "Sugars", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) 
  return(list(plot = ridge_plot, data = filtered_data))
}

# Generic function to create sample count tables
create_sample_count_table <- function(filtered_data, plant_name) {
  col_name <- paste(plant_name, "samples")
  
  counts <- filtered_data %>%
    select(filename, method) %>%
    distinct() %>%
    count(method) %>%
    mutate(!!col_name := paste(n, "&nbsp;&nbsp;", method)) %>%
    select(all_of(col_name))
  
  return(counts)
}

# Create plots and tables for each plant
plants <- sort(unique(dpt_data$plant))

for (pt in plants) {
  # Create plot and get filtered data
  result <- create_plant_faceted_plot(dpt_data, pt)
  
  # Print the plot
  print(result$plot)
  
  # Create and print sample count table with results='asis'
  counts_table <- create_sample_count_table(result$data, pt)
  cat(knitr::kable(counts_table, format = "html", escape = FALSE))
  cat("<br><br>")  # Add HTML line breaks for spacing
}
```

### Create helper function for stacked range plots
```{r stacked-range-plot-function}
# Builds an "average + range" stacked overlay plot for an arbitrary set of plants.
# data: dpt_data (or similar), filtered internally to plants_ordered
# plants: character vector of plant names, bottom to top in the final plot

build_stacked_range_plot <- function(data, plants, plant_palette, offset_amount = 1.67) {

  plot_data <- data %>%
    filter(plant %in% plants) %>%
    mutate(plant = factor(plant, levels = plants)) %>%
    mutate(snv_corrected = as.numeric(snv_corrected))

  # Calculate vertical offset for each plant to create stacking effect
  plant_offsets <- setNames(seq(0, length(plants) - 1) * offset_amount, plants)

  # Add offset to absorbance values based on plant
  plot_data <- plot_data %>%
    mutate(offset_absorbance = snv_corrected + plant_offsets[as.character(plant)])

  # Determine max y position for annotations
  max_y <- max(plot_data$offset_absorbance, na.rm = TRUE)

  # Create the base plot with functional region annotations
  base_plot <- ggplot() +
    scale_x_reverse(
      limits = c(3750, 650),
      breaks = seq(4000, 600, -500),  # Ticks every 500 from 4000 down to 600
      expand = c(0.001, 0.001)  # Remove padding on x-axis
    ) +

    labs(x = expression("Wavenumber (cm"^-1*")"),
         y = "Normalized absorbance") +
    theme(
      axis.text.y = element_blank(),
      axis.title.y = element_text(size = 15, face = "bold"),
      axis.text.x = element_text(size = 15, face = "bold"),
      axis.title.x = element_text(size = 16, face = "bold"),
      legend.position = "right",
      legend.justification = c(0, 0.01),  # (x, y) position; 0,0 is bottom-right
      legend.box = "vertical",
      legend.spacing.y = unit(1.8, "cm"),
      legend.title = element_text(size = 15, face = "bold"),
      legend.text = element_text(size = 14.5),
      plot.margin = margin(7, 14, 7, 7, "pt"),  # (top, right, bottom, left)
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank()
    ) +
    scale_y_continuous(
      breaks = plant_offsets + 0.003, # y-tick positions
      labels = NULL,
      expand = expansion(mult = c(0.003, 0.015)) # (bottom padding, top padding)
    ) +
    scale_x_reverse(
      limits = c(3750, 650),
      breaks = seq(3500, 600, -500),  # Ticks every 500 from 3500 down to 500
      expand = c(0.002, 0.003)  # Remove padding on x-axis
    ) +
    # Add shaded regions with annotations
     # Add simple plant region annotation
    annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("text", x = 2900, y = max_y * 0.7, label = "Simple Plant", 
             size = 3.6, fontface = "bold", hjust = 0.5, color = annotation_text_color) +
    # Add carbonyl region annotation
    annotate("rect", xmin = 1780, xmax = 1690, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#83a0e8") +
    annotate("text", x = 1940, y = max_y * 0.5, label = "Carbonyl", 
             size = 3.6, fontface = "bold", hjust = 0.5, color = annotation_text_color) +
    # Add microbial region annotation
    annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#a8cbad") +
    annotate("text", x = 1640, y = max_y * 0.67, label = "Microbial", 
             size = 3.6, fontface = "bold", hjust = 0.5, color = annotation_text_color) +
    # Add complex plant region annotation
    annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#beb5d5") +
    annotate("text", x = 1300, y = max_y * 0.80, label = "Complex Plant", 
             size = 3.6, fontface = "bold", hjust = 0.5, color = annotation_text_color) +
    # Add sugars region annotation
    annotate("rect", xmin = 1200, xmax = 900, ymin = -Inf, ymax = Inf, 
             alpha = 0.17, fill = "#c9aac7") +
    annotate("text", x = 911, y = max_y * 0.98, label = "Sugars", 
             size = 3.6, fontface = "bold", hjust = 0.5, color = annotation_text_color) 
  # Add layers for each plant with average and range
  combined_plot <- base_plot

  for (current_plant in plants) {
    plant_data <- plot_data %>% filter(plant == current_plant)

    for (mt in unique(plant_data$method)) {
      mt_data <- plant_data %>% filter(method == mt)

      # Calculate average and range for this plant+method combination
      summary_data <- mt_data %>%
        group_by(wavenumber) %>%
        summarise(
          mean_absorbance = mean(offset_absorbance, na.rm = TRUE),
          min_absorbance = min(offset_absorbance, na.rm = TRUE),
          max_absorbance = max(offset_absorbance, na.rm = TRUE),
          .groups = "drop"
        )

      # Add range ribbon (semi-transparent)
      combined_plot <- combined_plot +
        geom_ribbon(data = summary_data,
                    aes(x = wavenumber, ymin = min_absorbance, ymax = max_absorbance),
                    fill = plant_palette[paste(current_plant, mt, sep = "_")],
                    alpha = 0.58,
                    color = NA,
                    show.legend = FALSE)

      # Add average line (opaque)
      combined_plot <- combined_plot +
        geom_line(data = summary_data,
                  aes(x = wavenumber, y = mean_absorbance),
                  color = plant_palette[paste(current_plant, mt, sep = "_")],
                  linewidth = 0.88,
                  alpha = 1,
                  show.legend = FALSE)
    }
  }

  # Build one legend per plant using invisible dummy layers + ggnewscale, so each
  # plant keeps its own color palette in its own legend block.
  # plants goes bottom-to-top in the actual plot (first element = bottom offset),
  # so the legend blocks are built in the REVERSE order to match: the plant at the top of
  # the plot gets the top legend block, working down to the plant at the bottom.
  if (!require(ggnewscale, quietly = TRUE)) {
    install.packages("ggnewscale")
    library(ggnewscale)
  }

  legend_order_plants <- rev(plants)

  for (i in seq_along(legend_order_plants)) {
    current_plant <- legend_order_plants[i]

    # Only include methods that actually have data for this plant
    plant_plot_data <- plot_data %>%
    filter(plant == current_plant)
    present_mts <- sort(unique(plant_plot_data$method))
    if (length(present_mts) == 0) next

    dummy_data <- data.frame(
      x = rep(3777, length(present_mts)),
      y = rep(0, length(present_mts)),
      mt = present_mts
    )

    combined_plot <- combined_plot +
      geom_line(data = dummy_data, aes(x = x, y = y, color = mt), linewidth = 0, alpha = 0) +
      scale_color_manual(
        name = str_to_title(current_plant),
        values =     setNames(
        plant_palette[paste(current_plant,
                            present_mts,
                            sep="_")],
        present_mts
    ),
        breaks = present_mts,
        labels = present_mts,
        guide = guide_legend(override.aes = list(linewidth = 3, alpha = 1), order = i)
      )

    if (i < length(legend_order_plants)) {
      combined_plot <- combined_plot + ggnewscale::new_scale_color()
    }
  }

  return(list(plot = combined_plot, data = plot_data))
}
```

### Plot all plants \& methods
```{r stacked-range-plot-all-plants}
stacked_plot <- build_stacked_range_plot(
  data = dpt_data, 
  plants = plants, 
  plant_palette = plant_palette
  )

print(stacked_plot$plot)
ggsave_themed("combined_overlay_range_all_plants.png", stacked_plot$plot, width = 13, height = 9, dpi = 300)
```
# Integration of functional regions
Modified from "Berhe_Ghezzehei_Lab/DRIFTS Data Analysis/Visualizing_FTIR_w_Elemental_Data.R" which starts by calculating aliphatic and aromatic indices, then deriving simple and complex plant and microbial proportions before visualizing everything. This procedure sums non-normalized absorbance values over a specified range, and divides by the total peak area in the ranges of interest to get the proportions.

## Performing the integration
```{r spectral-integration-function}
# Function to perform spectral integration on DPT data (based on Leila_code_modified.Rmd)
perform_spectral_integration <- function(dpt_data) {
  cat("Starting spectral integration processing...\n")
  
  # Load required library for integration
  if (!require(pracma, quietly = TRUE)) {
    install.packages("pracma")
    library(pracma)
  }
  
  # Prepare the data in the format needed for integration
  # Rename columns to match expected format
  comb <- dpt_data %>%
    select(filename, wavenumber, absorbance) %>%
    dplyr::rename(sample = filename, abs = absorbance)
  
  # Define spectral windows of interest (same as Leila's code)
  red1 <- comb[comb$wavenumber > 2839 & comb$wavenumber < 2870, ]  # Aliphatic 1
  red2 <- comb[comb$wavenumber > 2898 & comb$wavenumber < 2976, ]  # Aliphatic 2  
  red3 <- comb[comb$wavenumber > 1690 & comb$wavenumber < 1780, ]  # Carbonyl
  red4 <- comb[comb$wavenumber > 1580 & comb$wavenumber < 1660, ]  # Microbial
  red5 <- comb[comb$wavenumber > 1500 & comb$wavenumber < 1550, ]  # Complex plant
  red6 <- comb[comb$wavenumber > 900 & comb$wavenumber < 1200, ]   # Sugars
  
  # Get unique samples
  sample <- unique(comb$sample)
  
  # Create integration results data frame
  ints <- data.frame(
    sample = sample,
    int_2839_2870 = numeric(length(sample)),
    int_2898_2976 = numeric(length(sample)),
    int_1690_1780 = numeric(length(sample)),
    int_1580_1660 = numeric(length(sample)),
    int_1500_1550 = numeric(length(sample)),
    int_900_1200 = numeric(length(sample)),
    stringsAsFactors = FALSE
  )
  
  # Calculate integrated areas for each spectral window and sample
  cat("Processing", length(sample), "samples for integration...\n")
  for (i in seq_len(nrow(ints))) {
    current_sample <- sample[i]
    
    # Extract data for current sample from each spectral window
    sample_red1 <- red1[red1$sample == current_sample, ]
    sample_red2 <- red2[red2$sample == current_sample, ]
    sample_red3 <- red3[red3$sample == current_sample, ]
    sample_red4 <- red4[red4$sample == current_sample, ]
    sample_red5 <- red5[red5$sample == current_sample, ]
    sample_red6 <- red6[red6$sample == current_sample, ]
    
    # Calculate area under curve (integration) using trapezoidal rule
    if (nrow(sample_red1) > 1) {
      ints$int_2839_2870[i] <- pracma::trapz(sample_red1$wavenumber, sample_red1$abs)
    }
    
    if (nrow(sample_red2) > 1) {
      ints$int_2898_2976[i] <- pracma::trapz(sample_red2$wavenumber, sample_red2$abs)
    }
    
    if (nrow(sample_red3) > 1) {
      ints$int_1690_1780[i] <- pracma::trapz(sample_red3$wavenumber, sample_red3$abs)
    }
    
    if (nrow(sample_red4) > 1) {
      ints$int_1580_1660[i] <- pracma::trapz(sample_red4$wavenumber, sample_red4$abs)
    }
    
    if (nrow(sample_red5) > 1) {
      ints$int_1500_1550[i] <- pracma::trapz(sample_red5$wavenumber, sample_red5$abs)
    }
    
    if (nrow(sample_red6) > 1) {
      ints$int_900_1200[i] <- pracma::trapz(sample_red6$wavenumber, sample_red6$abs)
    }
  }
  
  # Use existing metadata from the input data (no need to re-extract)
  cat("Using existing metadata from DPT data...\n")
  basic_meta <- dpt_data %>%
    select(filename, plant, method, run_date, run_rep) %>%
    distinct() %>%
    dplyr::rename(sample = filename)
  
  # Combine metadata with integration results
  ftir_data <- merge(basic_meta, ints, by = "sample", all = TRUE)
  
  # Rename integration columns to match expected names
  names(ftir_data)[names(ftir_data) == "int_1580_1660"] <- "int_microbial"
  names(ftir_data)[names(ftir_data) == "int_1500_1550"] <- "int_complex_plant"
  names(ftir_data)[names(ftir_data) == "int_1690_1780"] <- "int_carbonyl"
  names(ftir_data)[names(ftir_data) == "int_900_1200"] <- "int_sugars"
  
  # Calculate derived metrics
  ftir_data$aliphatic <- ftir_data$int_2839_2870 + ftir_data$int_2898_2976
  ftir_data$total_peak_area <- ftir_data$aliphatic + ftir_data$int_microbial + ftir_data$int_complex_plant + ftir_data$int_carbonyl + ftir_data$int_sugars
  ftir_data$aliphatic_prop <- ftir_data$aliphatic / ftir_data$total_peak_area
  ftir_data$complex_plant_prop <- ftir_data$int_complex_plant / ftir_data$total_peak_area
  ftir_data$carbonyl_prop <- ftir_data$int_carbonyl / ftir_data$total_peak_area
  ftir_data$sugars_prop <- ftir_data$int_sugars / ftir_data$total_peak_area
  ftir_data$microbial_prop <- ftir_data$int_microbial / ftir_data$total_peak_area
  ftir_data$aliphatic_microbial_prop <- ftir_data$aliphatic_prop / ftir_data$microbial_prop
  ftir_data$complex_microbial_prop <- ftir_data$complex_plant_prop / ftir_data$microbial_prop
  
  # Remove individual aliphatic integration columns (keep only derived metrics)
  ftir_data <- ftir_data %>%
    select(-int_2839_2870, -int_2898_2976)
  
  # Save the processed data to the output directory
  today <- format(Sys.Date(), "%Y-%m-%d")
  output_file <- file.path(outputs_folder, paste0("processed_int_data_", today, ".csv"))
  write.csv(ftir_data, file = output_file, row.names = FALSE)
  cat("Saved processed integration data to:", output_file, "\n")
  
  cat("Integration processing completed successfully.\n")
  return(ftir_data)
}
# Try to find the most recent processed integration data file
ftir_file <- find_most_recent_processed_file(outputs_folder)

if (!is.null(ftir_file) && file.exists(ftir_file)) {
  ftir_data <- read_csv(ftir_file, show_col_types = FALSE)
  cat("Successfully loaded FTIR integration data from:", basename(ftir_file), "\n")
  cat("Dimensions:", dim(ftir_data), "\n")
  cat("Columns:", names(ftir_data), "\n")
} else {
  cat("No processed integration data found. Processing raw DPT data to generate integration results...\n")
  
  # Check if we have the required dpt_data_roots from previous chunk
  if (!exists("dpt_data") || nrow(dpt_data) == 0) {
    stop("No DPT data available for integration. Please ensure the DPT data import was successful.")
  }
  
  # Process raw DPT data using the integrated function
  ftir_data <- perform_spectral_integration(dpt_data)
  
  # Verify the processing was successful
  if (is.null(ftir_data) || nrow(ftir_data) == 0) {
    stop("Failed to process DPT data for integration.")
  }
  
  cat("Successfully processed", nrow(ftir_data), "samples from DPT data\n")
  cat("Generated columns:", names(ftir_data), "\n")
}
```

## Generating statistics
This code chunk calculates summary statistics (mean and standard error) for the derived metrics (aliphatic, carbonyl, microbial, complex plant, sugars proportions) grouped by plant, preparation method, and plant-method combinations. The results are stored in separate data frames for later visualization.

```{r integration-statistics}
# Calculate summary statistics by group
plant_stats <- ftir_data %>%
  mutate(plant = factor(plant, levels = sort(unique(plant)))) %>%
  group_by(plant) %>%
  summarise(
    Mean_aliphatic_prop = mean(aliphatic_prop, na.rm = TRUE),
    SE_aliphatic_prop = sd(aliphatic_prop, na.rm = TRUE) / sqrt(n()),
    Mean_carbonyl_prop = mean(carbonyl_prop, na.rm = TRUE),
    SE_carbonyl_prop = sd(carbonyl_prop, na.rm = TRUE) / sqrt(n()),
    Mean_microbial_prop = mean(microbial_prop, na.rm = TRUE),
    SE_microbial_prop = sd(microbial_prop, na.rm = TRUE) / sqrt(n()),
    Mean_complex_plant_prop = mean(complex_plant_prop, na.rm = TRUE),
    SE_complex_plant_prop = sd(complex_plant_prop, na.rm = TRUE) / sqrt(n()),
    Mean_sugars_prop = mean(sugars_prop, na.rm = TRUE),
    SE_sugars_prop = sd(sugars_prop, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

method_stats <- ftir_data %>%
  group_by(method) %>%
  summarise(
    Mean_aliphatic_prop = mean(aliphatic_prop, na.rm = TRUE),
    SE_aliphatic_prop = sd(aliphatic_prop, na.rm = TRUE) / sqrt(n()),
    Mean_carbonyl_prop = mean(carbonyl_prop, na.rm = TRUE),
    SE_carbonyl_prop = sd(carbonyl_prop, na.rm = TRUE) / sqrt(n()),
    Mean_microbial_prop = mean(microbial_prop, na.rm = TRUE),
    SE_microbial_prop = sd(microbial_prop, na.rm = TRUE) / sqrt(n()),
    Mean_complex_plant_prop = mean(complex_plant_prop, na.rm = TRUE),
    SE_complex_plant_prop = sd(complex_plant_prop, na.rm = TRUE) / sqrt(n()),
    Mean_sugars_prop = mean(sugars_prop, na.rm = TRUE),
    SE_sugars_prop = sd(sugars_prop, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

plant_method_stats <- ftir_data %>%
  mutate(plant = factor(plant, levels = sort(unique(plant)))) %>%
  group_by(plant, method) %>%
  summarise(
    Mean_aliphatic_prop = mean(aliphatic_prop, na.rm = TRUE),
    SE_aliphatic_prop = sd(aliphatic_prop, na.rm = TRUE) / sqrt(n()),
    Mean_carbonyl_prop = mean(carbonyl_prop, na.rm = TRUE),
    SE_carbonyl_prop = sd(carbonyl_prop, na.rm = TRUE) / sqrt(n()),
    Mean_microbial_prop = mean(microbial_prop, na.rm = TRUE),
    SE_microbial_prop = sd(microbial_prop, na.rm = TRUE) / sqrt(n()),
    Mean_complex_plant_prop = mean(complex_plant_prop, na.rm = TRUE),
    SE_complex_plant_prop = sd(complex_plant_prop, na.rm = TRUE) / sqrt(n()),
    Mean_sugars_prop = mean(sugars_prop, na.rm = TRUE),
    SE_sugars_prop = sd(sugars_prop, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )
```

# Visualizing functional group proportions by method
Percentage stacked bar charts comparing preparation methods within each plant. Each bar is one method; segments are the five functional-group region proportions (aliphatic/simple plant, carbonyl, microbial, complex plant, sugars) from `plant_method_stats`. Colors come from `get_method_region_palette()`: each method keeps its own fixed hue (from `method_colors`/`method_hues` in the config), and region is encoded as a lightness shade within that hue, ordered so adjacent segments in the stack are never two similar shades.

```{r method-region-bar-function}
# Region display/stacking order -- bottom to top. This order also drives the
# color assignment in get_method_region_palette(), so change it here only.
region_order <- c("aliphatic", "carbonyl", "microbial", "complex", "sugar")
region_labels <- c(
  aliphatic = "Simple Plant",
  carbonyl  = "Carbonyl",
  microbial = "Microbial",
  complex   = "Complex Plant",
  sugar     = "Sugars"
)

# Builds one plant's % stacked bar chart, one bar per method.
build_method_region_bar_plot <- function(stats_data, current_plant, title = NULL) {
  plant_data <- stats_data %>%
    filter(plant == current_plant)

  present_methods <- sort(unique(plant_data$method))

  if (is.null(title)) {
    title <- str_to_title(current_plant)
  }

  long_data <- plant_data %>%
    select(method,
           aliphatic = Mean_aliphatic_prop,
           carbonyl  = Mean_carbonyl_prop,
           microbial = Mean_microbial_prop,
           complex   = Mean_complex_plant_prop,
           sugar     = Mean_sugars_prop) %>%
    pivot_longer(cols = -method, names_to = "region", values_to = "prop") %>%
    mutate(
      region   = factor(region, levels = region_order),
      method   = factor(method, levels = present_methods),
      pct      = prop * 100,
      fill_key = paste(method, region, sep = "_")
    )

  region_palette <- get_method_region_palette(regions = region_order, methods = present_methods)

  ggplot(long_data, aes(x = method, y = prop, fill = fill_key, group = region)) +
    geom_col(position = position_stack(reverse = TRUE),
             linewidth = 0.00, width = 0.88) +
    geom_text(aes(label = paste0(round(pct, 1), "%")),
              position = position_stack(reverse = TRUE, vjust = 0.5),
              color = ifelse(long_data$region == "aliphatic","white", ifelse(long_data$region == "microbial", "white", "black")),
              size = 3.6, fontface = "bold") +
    scale_fill_manual(values = region_palette) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0.01, 0.02))) +
    guides(fill = "none") +
    labs(x = NULL, y = "Functional Group Proportion (%)", title = title) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
}
```

Because region is encoded only as lightness (not hue), one grayscale key explains the shading for every method/plant panel -- there's no need for a separate legend per panel.

```{r method-region-bar-plots, fig.width=4, fig.height=6}
plants_for_bar <- sort(unique(plant_method_stats$plant))

region_key_plot <- build_region_key(region_order, region_labels)

bar_plots <- lapply(plants_for_bar, function(pt) {
  build_method_region_bar_plot(plant_method_stats, pt)
})
names(bar_plots) <- plants_for_bar

for (pt in plants_for_bar) {
  print(bar_plots[[pt]])
  ggsave_themed(paste0("method_region_stackedbar_", pt, ".png"),
                bar_plots[[pt]], width = 4, height = 6, dpi = 300)
}

print(region_key_plot)
```

### Combined figure (all plants side-by-side)
```{r method-region-bar-combined, fig.width=13, fig.height=6}
combined_bar_plot <- cowplot::plot_grid(
  plotlist = c(bar_plots, list(region_key_plot)),
  nrow = 1,
  rel_widths = c(rep(1, length(bar_plots)), 0.45)
)

print(combined_bar_plot)
ggsave_themed("method_region_stackedbar_all_plants.png", combined_bar_plot,
              width = 13, height = 6, dpi = 300)
```
