---
title: "ATR Analysis - ITQB"
author: "Adrianne Seiden"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: darkly
    highlight: zenburn
    code_folding: hide
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding,
                     output_dir = "../adrianne-code/") })
---
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  results = "markup",
  include = TRUE
)
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))
## Installing packages ##
list_of_packages <- c(
  "plyr",
  "dplyr",
  "stringr",
  "reshape",
  "data.table",
  "nc",
  "readr",
  "tidyr",
  "pracma"
)

# install missing packages
new_packages <- list_of_packages[!(
  list_of_packages %in% installed.packages()[, "Package"]
)]
if (length(new_packages)) install.packages(new_packages)

# load packages
library(plyr)
library(dplyr)
library(stringr)
library(reshape)
library(data.table)
library(nc)
library(readr)
library(tidyr)
library(pracma)
library(ggplot2)
library(ggridges)
library(cowplot)
library(grid)
library(ggsignif)

# Set paths
dpt_folder <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/all_abs_DPT_files"
outputs_folder <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/ATR_outputs"


# Helper function to find the most recent processed integration data file
find_most_recent_processed_file <- function(data_dir) {
  pattern <- "processed_int_data_.*\\.csv$"
  files <- list.files(data_dir, pattern = pattern, full.names = TRUE)
  
  if (length(files) == 0) {
    return(NULL)
  }
  
  # Extract dates from filenames and find the most recent
  dates <- stringr::str_extract(basename(files), "\\d{4}-\\d{2}-\\d{2}")
  most_recent_idx <- which.max(as.Date(dates))
  
  cat("Found", length(files), "processed integration files\n")
  cat("Using most recent file:", basename(files[most_recent_idx]), "\n")
  
  return(files[most_recent_idx])
}

# Source centralized theme and color configuration
source(file.path(dirname(knitr::current_input()), "config/ITQB_color_config.R"))

```

# Overview
placeholder

# Preparing data for analysis

## Importing DPT data
This code is modified from code by Stephany Soledad Chacon, "Box> Salk Institute Project> Salk Data> FTIR_Intact_decomposition_pots> FTIR_analysis.Rmd". Using data in the DPT format, we can extrapolate sample attributes from the file names, then normalize and visualize spectra across replicates, crops, and treatments. The width of the line bounding the spectra represents the range of the data.
```{r import-dpt-data}
# Check if folder exists and list files
if (!dir.exists(dpt_folder)) {
  stop("Directory does not exist: ", dpt_folder)
}

# List and sort files
cat("Files found in directory:\n")
print(list.files(dpt_folder))
DPTfiles <- sort(list.files(dpt_folder, pattern = ".dpt"))
cat("Number of .dpt files:", length(DPTfiles), "\n")

# Read files individually and combine
dpt_data_list <- list()

for (i in seq_along(DPTfiles)) {
  file_path <- file.path(dpt_folder, DPTfiles[i])
  
  # Read as lines and parse manually (DPT files are space-delimited)
  lines <- readLines(file_path, warn = FALSE)
  lines <- lines[lines != ""]  # Remove empty lines
  
  # Split each line by whitespace
  split_lines <- strsplit(lines, "\\s+")
  
  # Convert to data frame
  temp_data <- tryCatch({
    data.frame(
      wavenumber = as.numeric(sapply(split_lines, `[`, 1)),
      absorbance = as.numeric(sapply(split_lines, `[`, 2)),
      stringsAsFactors = FALSE
    )
  }, error = function(e) NULL)
  
  if (!is.null(temp_data)) {
    # Add filename column
    temp_data$filename <- DPTfiles[i]
    
    # Store in list
    dpt_data_list[[i]] <- temp_data
  }
}

# Remove NULL entries (failed reads)
dpt_data_list <- dpt_data_list[!sapply(dpt_data_list, is.null)]

# Combine all data
dpt_data <- bind_rows(dpt_data_list)

# Remove any rows with NA values
dpt_data <- dpt_data[complete.cases(dpt_data), ]

# Extract metadata from filenames

# Helper to extract fields from filenames with the pattern
# AKS_suberin_[plant]_[washing-treatment]_[yymmdd]-[run].dpt
extract_filename_fields <- function(filename) {
  stem <- tools::file_path_sans_ext(basename(filename))
  matches <- stringr::str_match(
    stem,
    "^AKS_suberin_([^_]+)_([^_]+)_([0-9]{6})-([0-9]+)$"
  )

  if (is.na(matches[1, 1])) {
    return(tibble(
      plant = NA_character_,
      method = NA_character_,
      run_date = NA_character_,
      run_rep = NA_integer_
    ))
  }

  tibble(
    plant = tolower(matches[1, 2]),
    method = tolower(matches[1, 3]),
    run_date = as.Date(matches[1, 4], format = "%y%m%d"),
    run_rep = as.integer(matches[1, 5])
  )
}

file_metadata <- bind_rows(lapply(dpt_data$filename, extract_filename_fields))
dpt_data <- bind_cols(dpt_data, file_metadata)

# save as csv for record-keeping
write.csv(dpt_data, file.path(outputs_folder, "dpt_data.csv"), row.names = FALSE)


cat("Samples by plant:\n")
print(table( dpt_data$plant))
cat("Samples by treatment:\n") 
print(table( dpt_data$method))

```

### Preprocessing
Define color palettes for each plant and method combination. This will be used in the plots to maintain consistent coloring across different visualizations.
```{r preprocessing}
# define color palettes for each plant and method combination
methods <- sort(unique( dpt_data$method))
plants <- sort(unique( dpt_data$plant))

plant_palette <- get_plant_palette(
  methods = methods,
  plants = plants
)

dpt_data$color_key <-
    paste(dpt_data$plant,
          dpt_data$method,
          sep = "_")

plant_colors <- get_plant_colors(sort(unique(dpt_data$plant)))

```

## Normalizing data
This code is modified to use baseline correction (subtracting the minimum) and standard normal variate (SNV) normalization.
```{r normalize-data}
# Normalize absorbance data by sample (same approach as OPUSprocessing.Rmd)
dpt_data <- dpt_data %>%
  group_by(filename) %>%
  mutate(
    # Baseline correction - subtract minimum
    baseline_corrected = absorbance - min(absorbance, na.rm = TRUE),
    # Apply SNV (Standard Normal Variate) normalization
    mean_abs = mean(baseline_corrected, na.rm = TRUE),
    sd_abs = sd(baseline_corrected, na.rm = TRUE),
    normalized_absorbance = (baseline_corrected - mean_abs) / sd_abs
  ) %>%
  select(-mean_abs, -sd_abs) %>%
  ungroup()
dpt_data <- dpt_data %>%
  group_by(filename) %>%
  mutate(
    # Baseline correction - subtract minimum
    snv_corrected = normalized_absorbance - min(normalized_absorbance, na.rm = TRUE)
  ) %>%
  ungroup()

dpt_data <- dpt_data %>%
  group_by(filename) %>%
  mutate(
    # Baseline correction + 1
    snv_plusOne = normalized_absorbance + 1
  ) %>%
  ungroup()

## Save normalized data for record-keeping

write.csv(dpt_data, file.path(outputs_folder, "dpt_data_normalized.csv"), row.names = FALSE)
```


# Plotting ATR spectra
## Plotting by preparation method
### Ridge plots
These plots are of normalized absorbance values. The thickness of the opaque line represents the range of the data. Annotations indicate integration regions used for calculations of simple plant, complex plant, and microbial organic matter proportions.
Integration regions according to Mainka et al. (2022):

Mainka, M., Summerauer, L., Wasner, D., Garland, G., Griepentrog, M., Berhe, A. A., & Doetterl, S. (2022). Soil geochemistry as a driver of soil organic matter composition: Insights from a soil chronosequence. Biogeosciences, 19(6), 1675–1689. https://doi.org/10.5194/bg-19-1675-2022

Plus carbonyl region (1800 - 1700 cm<sup>-1</sup>) added to the analysis.

<table style="width: auto; border-collapse: collapse; border: 0.1px solid #f5f5f5;">
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Simple plant (aliphatic):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">2976 - 2898 cm<sup>-1</sup> + 2870 - 2839 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Microbial:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1660 - 1580 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Complex plant (aromatic):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1550 - 1500 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Carbonyl:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1800 - 1700 cm<sup>-1</sup></td>
</tr>
</table>

```{r prep-method-plots, results='asis'}

# Generic function to create ridge plots for any preparation method
create_method_ridge_plot <- function(data, target_method, title = NULL) {
  # Filter data for specific method
  filtered_data <- data %>%
    filter(method == target_method)

  # Create title if not provided
  if (is.null(title)) {
    title <- paste("Roots (", target_method, ")", sep = "")
  }
  
  # Determine y-position for annotations based on data
  present_levels <- sort(unique(filtered_data$plant))  # only levels actually in this week's data, in original order
  peak_heights <- filtered_data %>%
    group_by(plant) %>%
    summarise(group_max = max(snv_plusOne, na.rm = TRUE)) %>%
    mutate(baseline = match(as.character(plant), present_levels))  # true rendered position, not raw factor code
  global_max <- max(peak_heights$group_max, na.rm = TRUE)
  max_y <- max(peak_heights$baseline + 3 * (peak_heights$group_max / global_max)) + 0.2
#   n_groups <- length(unique(filtered_data$plant))
#   ridge_scale <- 3  # matches your geom_density_ridges(scale = 3)
#   max_y <- n_groups + ridge_scale * .33 + 0.33  # headroom above the top category


  
  # Create the plot
  ridge_plot <- filtered_data %>%
    ggplot(mapping = aes(x = wavenumber,
                         y = plant,
                         height = snv_plusOne,
                         group = plant,
                         fill = plant,
                         color = plant)) +
    geom_density_ridges(stat = "identity",
                        scale = 3, # overlap scale - 1 is no overlap, >1 is more overlap
                        alpha = 0.25) +
    geom_vline(xintercept = 1500, linetype = "dotted", linewidth = 0.23) +
    geom_vline(xintercept = 1550, linetype = "dotted", linewidth = 0.23) +
    geom_vline(xintercept = 1580, linetype = "dashed", linewidth = 0.23) +
    geom_vline(xintercept = 1660, linetype = "dashed", linewidth = 0.23) +
    geom_vline(xintercept = 2839, linetype = "longdash", linewidth = 0.23) +
    geom_vline(xintercept = 2870, linetype = "longdash", linewidth = 0.23) +
    geom_vline(xintercept = 2898, linetype = "longdash", linewidth = 0.23) +
    geom_vline(xintercept = 2976, linetype = "longdash", linewidth = 0.23) +
    xlim(3777, 422) +
    scale_y_discrete(expand = expansion(mult = c(0, 0.1), add = c(0, 0))) +  # space below, space above+  # space below, space above
    
    theme(axis.text.y = element_text(size = 12),
          legend.position = "none") +
    scale_fill_manual(values = plant_colors) +
    scale_color_manual(values = plant_colors) +
    ggtitle(title) +
    labs(x = expression("wavenumber (cm"^-1*")"), y = "") +
    # Add simple plant region annotation
    annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("text", x = 2560, y = max_y, label = "Simple Plant", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add microbial region annotation
    annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#a8cbad") +
    annotate("text", x = 1870, y = max_y + 0.11, label = "Microbial", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add complex plant region annotation
    annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#beb5d5") +
    annotate("text", x = 1190, y = max_y + 0.11, label = "Complex Plant", 
             size = 3.3, hjust = 0.5, color = annotation_text_color)
  
  return(list(plot = ridge_plot, data = filtered_data))
}

# Generic function to create sample count tables
create_sample_count_table <- function(filtered_data, method) {
  col_name <- paste(method, "samples")
  
  counts <- filtered_data %>%
    select(filename, plant) %>%
    distinct() %>%
    count(plant) %>%
    mutate(!!col_name := paste(n, "&nbsp;&nbsp;", plant)) %>%
    select(all_of(col_name))
  
  return(counts)
}

# Create plots and tables for each method
methods <- sort(unique(dpt_data$method))

for (mt in methods) {
  # Create plot and get filtered data
  result <- create_method_ridge_plot(dpt_data, mt)
  
  # Print the plot
  print(result$plot)
  
  # Create and print sample count table with results='asis'
  counts_table <- create_sample_count_table(result$data, mt)
  cat(knitr::kable(counts_table, format = "html", escape = FALSE))
  cat("<br><br>")  # Add HTML line breaks for spacing
}
```

## Plotting by plant
These plots are of normalized absorbance values. The thickness of the opaque line represents the range of the data. Annotations indicate integration regions used for calculations of simple plant, complex plant, and microbial organic matter proportions.  

<table style="width: auto; border-collapse: collapse; border: 0.1px solid #f5f5f5;">
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Simple plant (aliphatic):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">2976 - 2898 cm<sup>-1</sup> + 2870 - 2839 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Microbial:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1660 - 1580 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Complex plant (aromatic):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1550 - 1500 cm<sup>-1</sup></td>
</tr>
</table>

### Ridge plots
```{r plant-plots, results='asis'}

# Generic function to create plant-specific ATR plots
create_plant_plot <- function(data, plant_name, color_palette) {
  # Filter data for specific plant
  plant_data <- data %>%
    group_by(method) %>%
    filter(plant == plant_name)
  
  # Ensure absorbance is numeric
  plant_data$snv_plusOne <- as.numeric(plant_data$snv_plusOne)
  plant_data$color_key <- paste(plant_data$plant, plant_data$method, sep = "_")
  # Create the plot
  plant_plot <- ggplot(plant_data,
                       aes(x = wavenumber,
                           y = snv_plusOne,
                           color = color_key)) +
    xlim(3777, 422) +
    ylim(0, max(plant_data$snv_plusOne, na.rm = TRUE) + 0.1) +
    geom_line() +
    
    # scale_color_manual(values = plant_palette) +
    ggtitle(paste(str_to_title(plant_name), "root ATR")) +
    labs(x = expression("wavenumber (cm"^-1*")"), 
         y = "normalized absorbance") +
    geom_vline(xintercept = 1500, linetype = "dotted", linewidth = 0.3) +
    geom_vline(xintercept = 1550, linetype = "dotted", linewidth = 0.3) +
    geom_vline(xintercept = 1580, linetype = "dashed", linewidth = 0.23) +
    geom_vline(xintercept = 1660, linetype = "dashed", linewidth = 0.23) +
    geom_vline(xintercept = 2839, linetype = "longdash", linewidth = 0.23) +
    geom_vline(xintercept = 2870, linetype = "longdash", linewidth = 0.23) +
    geom_vline(xintercept = 2898, linetype = "longdash", linewidth = 0.23) +
    geom_vline(xintercept = 2976, linetype = "longdash", linewidth = 0.23) +
    # Add simple plant region annotation
    annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("text", x = 2550, y = max(plant_data$snv_plusOne), label = "Simple Plant", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add microbial region annotation
    annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#a8cbad") +
    annotate("text", x = 1850, y = max(plant_data$snv_plusOne), label = "Microbial", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add complex plant region annotation
    annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#beb5d5") +
    annotate("text", x = 1200, y = max(plant_data$snv_plusOne)+0.02, label = "Complex Plant", 
             size = 3.3, hjust = 0.5, color = annotation_text_color)
  
  return(list(plot = plant_plot, data = plant_data))
}

# Generic function to create plant ID tables
create_plant_id_table <- function(plant_data, plant_name) {
  combinations <- plant_data %>%
    select(plant, method) %>%
    distinct() %>%
    arrange(plant, method) %>%
    group_by(method) %>%
    summarise(pot_ids = {
      ids <- paste(plant, sep = "")
      # Group every 3 items and join with line breaks
      grouped <- split(ids, ceiling(seq_along(ids)/5))
      lines <- sapply(grouped, function(x) paste(x, collapse = ",&ensp;"))
      paste(lines, collapse = "<br>")
    }, .groups = 'drop') %>%
    pivot_wider(names_from = method, values_from = pot_ids)
  
  return(combinations)
}

# Generate plots and tables for each plant
plants <- sort(unique( dpt_data$plant))

for (current_plant in plants) {
  # Create plot and get data
  result <- create_plant_plot(
    dpt_data,
    current_plant,
    plant_palette
)
  
  # Print the plot
  print(result$plot)
  cat("<br>")  # Add space between plot and table
  
  # Create and print ID combinations table
  id_table <- create_plant_id_table(result$data, current_plant)
  cat(paste0("<h4>", str_to_title(current_plant), " root samples</h4>"))
  cat(knitr::kable(id_table, format = "html", escape = FALSE, table.attr = 'style="border-collapse: separate; border-spacing: 35px 0;"'))
  cat("<br><br>")  # Add HTML line breaks for spacing
}
```

### Create helper function for stacked range plots
```{r stacked-range-plot-function}
# Builds an "average + range" stacked overlay plot for an arbitrary set of plants.
# data: dpt_data (or similar), filtered internally to plants_ordered
# plants: character vector of plant names, bottom to top in the final plot

build_stacked_range_plot <- function(data, plants, plant_palette, offset_amount = 1.67) {

  plot_data <- data %>%
    filter(plant %in% plants) %>%
    mutate(plant = factor(plant, levels = plants)) %>%
    mutate(snv_plusOne = as.numeric(snv_plusOne))

  # Calculate vertical offset for each plant to create stacking effect
  plant_offsets <- setNames(seq(0, length(plants) - 1) * offset_amount, plants)

  # Add offset to absorbance values based on plant
  plot_data <- plot_data %>%
    mutate(offset_absorbance = snv_plusOne + plant_offsets[as.character(plant)])

  # Determine max y position for annotations
  max_y <- max(plot_data$offset_absorbance, na.rm = TRUE)

  # Create the base plot with functional region annotations
  base_plot <- ggplot() +
    xlim(3777, 422) +

    labs(x = expression("Wavenumber (cm"^-1*")"),
         y = "Normalized absorbance") +
    theme(
      axis.text.y = element_blank(),
      axis.title.y = element_text(size = 15, face = "bold"),
      axis.text.x = element_text(size = 15, face = "bold"),
      axis.title.x = element_text(size = 16, face = "bold"),
      legend.position = "right",
      legend.justification = c(0, 0.01),  # (x, y) position; 0,0 is bottom-right
      legend.box = "vertical",
      legend.spacing.y = unit(1.8, "cm"),
      legend.title = element_text(size = 15, face = "bold"),
      legend.text = element_text(size = 14.5),
      plot.margin = margin(7, 14, 7, 7, "pt"),  # (top, right, bottom, left)
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank()
    ) +
    scale_y_continuous(
      breaks = plant_offsets + 0.003, # y-tick positions
      labels = NULL,
      expand = expansion(mult = c(0.003, 0.015)) # (bottom padding, top padding)
    ) +
    scale_x_reverse(
      limits = c(3770, 415),
      breaks = seq(3500, 500, -500),  # Ticks every 500 from 3500 down to 500
      expand = c(0.002, 0.003)  # Remove padding on x-axis
    ) +
    # Add vertical lines for functional regions
    geom_vline(xintercept = 1500, linetype = "dotted", linewidth = 0.46, color = annotation_text_color, alpha = 0.8) +
    geom_vline(xintercept = 1550, linetype = "dotted", linewidth = 0.46, color = annotation_text_color, alpha = 0.8) +
    geom_vline(xintercept = 1580, linetype = "dashed", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
    geom_vline(xintercept = 1660, linetype = "dashed", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
    geom_vline(xintercept = 2839, linetype = "longdash", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
    geom_vline(xintercept = 2870, linetype = "longdash", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
    geom_vline(xintercept = 2898, linetype = "longdash", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
    geom_vline(xintercept = 2976, linetype = "longdash", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
    # Add shaded regions with annotations
    annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf,
             alpha = 0.3, fill = "#e3c8b1") +
    annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf,
             alpha = 0.3, fill = "#e3c8b1") +
    annotate("text", x = 2650, y = max_y - 0.05, label = "Simple Plant",
             size = 4.9, fontface = "bold", hjust = 0.5, color = annotation_text_color) +
    annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf,
             alpha = 0.3, fill = "#a8cbad") +
    annotate("text", x = 1800, y = max_y, label = "Microbial",
             size = 4.9, fontface = "bold", hjust = 0.5, color = annotation_text_color) +
    annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf,
             alpha = 0.3, fill = "#beb5d5") +
    annotate("text", x = 1300, y = max_y, label = "Complex Plant",
             size = 4.9, fontface = "bold", hjust = 0.5, color = annotation_text_color)

  # Add layers for each plant with average and range
  combined_plot <- base_plot

  for (current_plant in plants) {
    plant_data <- plot_data %>% filter(plant == current_plant)

    for (mt in unique(plant_data$method)) {
      mt_data <- plant_data %>% filter(method == mt)

      # Calculate average and range for this plant+method combination
      summary_data <- mt_data %>%
        group_by(wavenumber) %>%
        summarise(
          mean_absorbance = mean(offset_absorbance, na.rm = TRUE),
          min_absorbance = min(offset_absorbance, na.rm = TRUE),
          max_absorbance = max(offset_absorbance, na.rm = TRUE),
          .groups = "drop"
        )

      # Add range ribbon (semi-transparent)
      combined_plot <- combined_plot +
        geom_ribbon(data = summary_data,
                    aes(x = wavenumber, ymin = min_absorbance, ymax = max_absorbance),
                    fill = plant_palette[paste(current_plant, mt, sep = "_")],
                    alpha = 0.58,
                    color = NA,
                    show.legend = FALSE)

      # Add average line (opaque)
      combined_plot <- combined_plot +
        geom_line(data = summary_data,
                  aes(x = wavenumber, y = mean_absorbance),
                  color = plant_palette[paste(current_plant, mt, sep = "_")],
                  linewidth = 0.88,
                  alpha = 1,
                  show.legend = FALSE)
    }
  }

  # Build one legend per plant using invisible dummy layers + ggnewscale, so each
  # plant keeps its own color palette in its own legend block.
  # plants goes bottom-to-top in the actual plot (first element = bottom offset),
  # so the legend blocks are built in the REVERSE order to match: the plant at the top of
  # the plot gets the top legend block, working down to the plant at the bottom.
  if (!require(ggnewscale, quietly = TRUE)) {
    install.packages("ggnewscale")
    library(ggnewscale)
  }

  legend_order_plants <- rev(plants)

  for (i in seq_along(legend_order_plants)) {
    current_plant <- legend_order_plants[i]

    # Only include methods that actually have data for this plant
    plant_plot_data <- plot_data %>%
    filter(plant == current_plant)
    present_mts <- sort(unique(plant_plot_data$method))
    if (length(present_mts) == 0) next

    dummy_data <- data.frame(
      x = rep(3777, length(present_mts)),
      y = rep(0, length(present_mts)),
      mt = present_mts
    )

    combined_plot <- combined_plot +
      geom_line(data = dummy_data, aes(x = x, y = y, color = mt), linewidth = 0, alpha = 0) +
      scale_color_manual(
        name = str_to_title(current_plant),
        values =     setNames(
        plant_palette[paste(current_plant,
                            present_mts,
                            sep="_")],
        present_mts
    ),
        breaks = present_mts,
        labels = present_mts,
        guide = guide_legend(override.aes = list(linewidth = 3, alpha = 1), order = i)
      )

    if (i < length(legend_order_plants)) {
      combined_plot <- combined_plot + ggnewscale::new_scale_color()
    }
  }

  return(list(plot = combined_plot, data = plot_data))
}
```

### Plot all plants
```{r stacked-range-plot-all-plants}
stacked_plot <- build_stacked_range_plot(
  data = dpt_data, 
  plants = plants, 
  plant_palette = plant_palette
  )

print(stacked_plot$plot)
ggsave_themed("combined_overlay_range_all_plants.png", stacked_plot$plot, width = 13, height = 9, dpi = 300)
```