---
title: "ATR Analysis - ITQB"
author: "Adrianne Seiden"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: darkly
    highlight: zenburn
    code_folding: hide
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding,
                     output_dir = "../adrianne-code/") })
---
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  results = "markup",
  include = TRUE
)
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))
## Installing packages ##
list_of_packages <- c(
  "plyr",
  "dplyr",
  "stringr",
  "reshape",
  "data.table",
  "nc",
  "readr",
  "tidyr",
  "pracma"
)

# install missing packages
new_packages <- list_of_packages[!(
  list_of_packages %in% installed.packages()[, "Package"]
)]
if (length(new_packages)) install.packages(new_packages)

# load packages
library(plyr)
library(dplyr)
library(stringr)
library(reshape)
library(data.table)
library(nc)
library(readr)
library(tidyr)
library(pracma)
library(ggplot2)
library(ggridges)
library(cowplot)
library(grid)
library(ggsignif)

# Set paths
dpt_folder <- "C:/Users/adria/OneDrive - University of California Merced/2026Summer/Portugal/Data/all_abs_DPT_files"
outputs_folder <- "C:/Users/adria/OneDrive - University of California Merced/2026Summer/Portugal/Data/ATR_outputs"


# Helper function to find the most recent processed integration data file
find_most_recent_processed_file <- function(data_dir) {
  pattern <- "processed_int_data_.*\\.csv$"
  files <- list.files(data_dir, pattern = pattern, full.names = TRUE)
  
  if (length(files) == 0) {
    return(NULL)
  }
  
  # Extract dates from filenames and find the most recent
  dates <- stringr::str_extract(basename(files), "\\d{4}-\\d{2}-\\d{2}")
  most_recent_idx <- which.max(as.Date(dates))
  
  cat("Found", length(files), "processed integration files\n")
  cat("Using most recent file:", basename(files[most_recent_idx]), "\n")
  
  return(files[most_recent_idx])
}

# Source centralized theme and color configuration
source(file.path(dirname(knitr::current_input()), "config/ITQB_color_config.R"))

```

# Overview
placeholder

# Preparing data for analysis

## Importing DPT data
This code is modified from code by Stephany Soledad Chacon, "Box> Salk Institute Project> Salk Data> FTIR_Intact_decomposition_pots> FTIR_analysis.Rmd". Using data in the DPT format, we can extrapolate sample attributes from the file names, then normalize and visualize spectra across replicates, crops, and treatments. The width of the line bounding the spectra represents the range of the data.
```{r import-dpt-data}
# Check if folder exists and list files
if (!dir.exists(dpt_folder)) {
  stop("Directory does not exist: ", dpt_folder)
}

# List and sort files
cat("Files found in directory:\n")
print(list.files(dpt_folder))
DPTfiles <- sort(list.files(dpt_folder, pattern = ".dpt"))
cat("Number of .dpt files:", length(DPTfiles), "\n")

# Read files individually and combine
dpt_data_list <- list()

for (i in seq_along(DPTfiles)) {
  file_path <- file.path(dpt_folder, DPTfiles[i])
  
  # Read as lines and parse manually (DPT files are space-delimited)
  lines <- readLines(file_path, warn = FALSE)
  lines <- lines[lines != ""]  # Remove empty lines
  
  # Split each line by whitespace
  split_lines <- strsplit(lines, "\\s+")
  
  # Convert to data frame
  temp_data <- tryCatch({
    data.frame(
      wavenumber = as.numeric(sapply(split_lines, `[`, 1)),
      absorbance = as.numeric(sapply(split_lines, `[`, 2)),
      stringsAsFactors = FALSE
    )
  }, error = function(e) NULL)
  
  if (!is.null(temp_data)) {
    # Add filename column
    temp_data$filename <- DPTfiles[i]
    
    # Store in list
    dpt_data_list[[i]] <- temp_data
  }
}

# Remove NULL entries (failed reads)
dpt_data_list <- dpt_data_list[!sapply(dpt_data_list, is.null)]

# Combine all data
dpt_data <- bind_rows(dpt_data_list)

# Remove any rows with NA values
dpt_data <- dpt_data[complete.cases(dpt_data), ]

# Extract metadata from filenames

# Helper to extract fields from filenames with the pattern
# AKS_suberin_[plant]_[washing-treatment]_[yymmdd]-[run].dpt
extract_filename_fields <- function(filename) {
  stem <- tools::file_path_sans_ext(basename(filename))
  matches <- stringr::str_match(
    stem,
    "^AKS_suberin_([^_]+)_([^_]+)_([0-9]{6})-([0-9]+)$"
  )

  if (is.na(matches[1, 1])) {
    return(tibble(
      plant = NA_character_,
      method = NA_character_,
      run_date = as.Date(NA),
      run_rep = NA_integer_
    ))
  }

  tibble(
    plant = tolower(matches[1, 2]),
    method = tolower(matches[1, 3]),
    run_date = as.Date(matches[1, 4], format = "%y%m%d"),
    run_rep = as.integer(matches[1, 5])
  )
}

# Check for problems in filename parsing
bad_files <- dpt_data$filename[
  !stringr::str_detect(
    basename(dpt_data$filename),
    "^AKS_suberin_[^_]+_[^_]+_[0-9]{6}-[0-9]+\\.dpt$"
  )
]

bad_files

file_metadata <- bind_rows(lapply(dpt_data$filename, extract_filename_fields))
dpt_data <- bind_cols(dpt_data, file_metadata)

# save as csv for record-keeping
write.csv(dpt_data, file.path(outputs_folder, "dpt_data.csv"), row.names = FALSE)


cat("Samples by plant:\n")
print(table( dpt_data$plant))
cat("Samples by treatment:\n") 
print(table( dpt_data$method))

```

### Preprocessing
Define color palettes for each plant and method combination. This will be used in the plots to maintain consistent coloring across different visualizations.
```{r preprocessing}
# define color palettes for each plant and method combination
methods <- sort(unique( dpt_data$method))
plants <- sort(unique( dpt_data$plant))

plant_palette <- get_plant_palette(
  methods = methods,
  plants = plants
)

dpt_data$color_key <-
    paste(dpt_data$plant,
          dpt_data$method,
          sep = "_")

plant_colors <- get_plant_colors(sort(unique(dpt_data$plant)))

```

## Normalizing data
This code is modified to use baseline correction (subtracting the minimum) and standard normal variate (SNV) normalization.
```{r normalize-data}
# Normalize absorbance data by sample (same approach as OPUSprocessing.Rmd)
dpt_data <- dpt_data %>%
  group_by(filename) %>%
  mutate(
    # Baseline correction - subtract minimum
    baseline_corrected = absorbance - min(absorbance, na.rm = TRUE),
    # Apply SNV (Standard Normal Variate) normalization
    mean_abs = mean(baseline_corrected, na.rm = TRUE),
    sd_abs = sd(baseline_corrected, na.rm = TRUE),
    normalized_absorbance = (baseline_corrected - mean_abs) / sd_abs
  ) %>%
  select(-mean_abs, -sd_abs) %>%
  ungroup()
dpt_data <- dpt_data %>%
  group_by(filename) %>%
  mutate(
    # Baseline correction - subtract minimum
    snv_corrected = normalized_absorbance - min(normalized_absorbance, na.rm = TRUE)
  ) %>%
  ungroup()

dpt_data <- dpt_data %>%
  mutate(
    material_class = unname(material_class_map[method]),
    extraction_bin = unname(extraction_bin_map[method]),
    material_class = factor(material_class, levels = c("raw", "insoluble", "soluble", "other")),
    extraction_bin = factor(extraction_bin, levels = c("wsol-1-2", "wsol-3-4", "wsol-5+", "wsol-pooled")),
    plot_group = as.character(material_class),
    plot_group = ifelse(material_class == "soluble", as.character(extraction_bin), plot_group),
    plot_group = factor(plot_group,
                         levels = c("raw", "insoluble", "wsol-1-2", "wsol-3-4", "wsol-5+", "wsol-pooled"))
  )

## Save normalized data for record-keeping

write.csv(dpt_data, file.path(outputs_folder, "dpt_data_normalized.csv"), row.names = FALSE)
```


# Plotting ATR spectra
## Plotting by preparation method
These plots are of normalized absorbance values. The thickness of the opaque line represents the range of the data. Annotations indicate integration regions used for calculations of simple plant, complex plant, carbonyl, protein/amide, and carbohydrate ("sugars") organic matter proportions.
Integration regions according to Mainka et al. (2022):

Mainka, M., Summerauer, L., Wasner, D., Garland, G., Griepentrog, M., Berhe, A. A., & Doetterl, S. (2022). Soil geochemistry as a driver of soil organic matter composition: Insights from a soil chronosequence. Biogeosciences, 19(6), 1675–1689. https://doi.org/10.5194/bg-19-1675-2022

Plus carbonyl region (1780 - 1690 cm<sup>-1</sup>) \& sugars region (1200 - 900 cm<sup>-1</sup>) added to the analysis.

<table style="width: auto; border-collapse: collapse; border: 0.1px solid #f5f5f5;">
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Aliphatics (simple plant):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">2976 - 2898 cm<sup>-1</sup> + 2870 - 2839 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Carbonyls:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1780 - 1690 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Protein/amide:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1660 - 1580 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Aromatics (complex plant):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1550 - 1500 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Sugars:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1200 - 900 cm<sup>-1</sup></td>
</tr>
</table>

```{r prep-method-plots, results='asis'}

# Generic function to create ridge plots for any preparation method
create_method_ridge_plot <- function(data, target_method, title = NULL) {
  # Filter data for specific method
  filtered_data <- data %>%
    filter(method == target_method)

  # Create title if not provided
  if (is.null(title)) {
    title <- paste("Roots (", target_method, ")", sep = "")
  }
  
  # Determine y-position for annotations based on data
  present_levels <- sort(unique(filtered_data$plant))  # only levels actually in this week's data, in original order

  max_y <- 7 - length(present_levels)  
  
  # Create the plot
  ridge_plot <- filtered_data %>%
    ggplot(mapping = aes(x = wavenumber,
                         y = plant,
                         height = snv_corrected,
                         group = plant,
                         fill = plant,
                         color = plant)) +
    geom_density_ridges(stat = "identity",
                        scale = 3, # overlap scale - 1 is no overlap, >1 is more overlap
                        alpha = 0.25) +
                        
    scale_x_reverse(
      limits = c(3750, 600),
      breaks = seq(3600, 600, -200),  # Ticks every 200 from 3600 down to 600
      expand = c(0.001, 0.001) 
      # Remove padding on x-axis
    ) +
    scale_y_discrete(expand = expansion(mult = c(0, 0.1), add = c(0, 0))) +  # space below, space above+  # space below, space above
    
    theme(axis.text.y = element_text(size = 12),
          legend.position = "none") +
    scale_fill_manual(values = plant_colors) +
    scale_color_manual(values = plant_colors) +
    ggtitle(title) +
    labs(x = expression("wavenumber (cm"^-1*")"), y = "") +
    # Add simple plant region annotation
    annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("text", x = 2900, y = max_y * 1.08, label = "Simple Plant", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add carbonyl region annotation
    annotate("rect", xmin = 1780, xmax = 1690, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#83a0e8") +
    annotate("text", x = 1910, y = max_y * 0.9, label = "Carbonyl", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add protein/amide region annotation
    annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#a8cbad") +
    annotate("text", x = 1640, y = max_y * 1.05, label = "Protein/amide", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add complex plant region annotation
    annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#beb5d5") +
    annotate("text", x = 1300, y = max_y * 1.22, label = "Complex Plant", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) +
    # Add sugars region annotation
    annotate("rect", xmin = 1200, xmax = 900, ymin = -Inf, ymax = Inf, 
             alpha = 0.17, fill = "#c9aac4") +
    annotate("text", x = 933, y = max_y * 1.4, label = "Sugars", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) 
  
  return(list(plot = ridge_plot, data = filtered_data))
}

# Generic function to create sample count tables
create_sample_count_table <- function(filtered_data, method) {
  col_name <- paste(method, "samples")
  
  counts <- filtered_data %>%
    select(filename, plant) %>%
    distinct() %>%
    count(plant) %>%
    mutate(!!col_name := paste(n, "&nbsp;&nbsp;", plant)) %>%
    select(all_of(col_name))
  
  return(counts)
}

# Create plots and tables for each method
methods <- sort(unique(dpt_data$method))

for (mt in methods) {
  # Create plot and get filtered data
  result <- create_method_ridge_plot(dpt_data, mt)
  
  # Print the plot
  print(result$plot)
  
  # Create and print sample count table with results='asis'
  counts_table <- create_sample_count_table(result$data, mt)
  cat(knitr::kable(counts_table, format = "html", escape = FALSE))
  cat("<br><br>")  # Add HTML line breaks for spacing
}

# troubleshooting method colors
method_levels <- sort(unique(dpt_data$method))

if (!all(method_levels %in% names(method_colors))) {
  missing_colors <- setdiff(method_levels, names(method_colors))
  cat(missing_colors," do not have defined colors in method_colors.")
}

```

## Plotting by plant
These plots are of normalized absorbance values. The thickness of the opaque line represents the range of the data. Annotations indicate integration regions used for calculations of simple plant, complex plant, and protein/amide organic matter proportions.  

<table style="width: auto; border-collapse: collapse; border: 0.1px solid #f5f5f5;">
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Aliphatics (simple plant):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">2976 - 2898 cm<sup>-1</sup> + 2870 - 2839 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Carbonyls:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1780 - 1690 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Protein/Amide:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1660 - 1580 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Aromatics (complex plant):</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1550 - 1500 cm<sup>-1</sup></td>
</tr>
<tr>
<td style="padding: 5px; white-space: nowrap; border: 0.1px dotted #f5f5f5;">Sugars:</td>
<td style="padding: 5px; border: 0.1px dotted #f5f5f5;">1200 - 900 cm<sup>-1</sup></td>
</tr>
</table>

```{r facet-plant-plots, results='asis'}
# check which methods do not have defined colors

# Generic function to create ridge plots for any preparation method
create_plant_faceted_plot <- function(data, current_plant, title = NULL) {
  # Filter data for specific method
  filtered_data <- data %>%
    filter(plant == current_plant)

  # Create title if not provided
  if (is.null(title)) {
    title <- paste(str_to_title(current_plant), "roots", sep = " ")
  }
  
  # Determine y-position for annotations based on data
  present_levels <- sort(unique(filtered_data$method))  # only levels actually in this week's data, in original order

  max_y <- 7 - length(present_levels)/5
  
  # Fixed method colors (from ITQB_color_config.R) -- method is now the hue,
  # not a shade of this plant's hue, so a given method is always the same
  # color here no matter which plant it's shown for.
  method_colors_current <- get_method_colors(present_levels)

  # Create the plot
  ridge_plot <- filtered_data %>%
    ggplot(mapping = aes(x = wavenumber,
                         y = method,
                         height = snv_corrected,
                         group = method,
                         fill = method,
                         color = method)) +
    geom_density_ridges(stat = "identity",
                        scale = 3, # overlap scale - 1 is no overlap, >1 is more overlap
                        alpha = 0.25) +
   scale_x_reverse(
      limits = c(3750, 600),
      breaks = seq(3600, 600, -200),  # Ticks every 200 from 3600 down to 600
      expand = c(0.001, 0.001) 
      # Remove padding on x-axis
    ) +
    scale_y_discrete(expand = expansion(mult = c(0, 0.1), add = c(0, 0))) +  # space below, space above+  # space below, space above
    
    theme(axis.text.y = element_text(size = 12),
          legend.position = "none") +
    scale_fill_manual(values = method_colors_current) +
    scale_color_manual(values = method_colors_current) +
    ggtitle(title) +
    labs(x = expression("wavenumber (cm"^-1*")"), y = "") +
    # Add simple plant region annotation
    annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#e3c8b1") +
    annotate("text", x = 2900, y = max_y*1.03, label = "Simple Plant", 
             size = 3.33, hjust = 0.5, color = annotation_text_color) +
    # Add carbonyl region annotation
    annotate("rect", xmin = 1780, xmax = 1690, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#83a0e8") +
    annotate("text", x = 1900, y = max_y*0.95, label = "Carbonyl", 
             size = 3.33, hjust = 0.5, color = annotation_text_color) +
    # Add protein/amide region annotation
    annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#a8cbad") +
    annotate("text", x = 1640, y = max_y * 1.07, label = "Protein/Amide", 
             size = 3.33, hjust = 0.5, color = annotation_text_color) +
    # Add complex plant region annotation
    annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf, 
             alpha = 0.2, fill = "#beb5d5") +
    annotate("text", x = 1333, y = max_y * 1.22, label = "Complex Plant", 
             size = 3.33, hjust = 0.5, color = annotation_text_color) +
      # Add sugars region annotation
    annotate("rect", xmin = 1200, xmax = 900, ymin = -Inf, ymax = Inf, 
             alpha = 0.16, fill = "#c3aac1") +
    annotate("text", x = 933, y = max_y * 1.29, label = "Sugars", 
             size = 3.3, hjust = 0.5, color = annotation_text_color) 
  return(list(plot = ridge_plot, data = filtered_data))
}

# Generic function to create sample count tables
create_sample_count_table <- function(filtered_data, plant_name) {
  col_name <- paste(plant_name, "samples")
  
  counts <- filtered_data %>%
    select(filename, method) %>%
    distinct() %>%
    count(method) %>%
    mutate(!!col_name := paste(n, "&nbsp;&nbsp;", method)) %>%
    select(all_of(col_name))
  
  return(counts)
}

# Create plots and tables for each plant
plants <- sort(unique(dpt_data$plant))

for (pt in plants) {
  # Create plot and get filtered data
  result <- create_plant_faceted_plot(dpt_data, pt)
  
  # Print the plot
  print(result$plot)
  
  # Create and print sample count table with results='asis'
  counts_table <- create_sample_count_table(result$data, pt)
  cat(knitr::kable(counts_table, format = "html", escape = FALSE))
  cat("<br><br>")  # Add HTML line breaks for spacing
}
```

### Create helper function for stacked range plots
```{r stacked-range-plot-function}
# Suberin-associated peak positions (wavenumber, cm^-1) and the color used to
# mark each one. Kept together as ONE data frame (rather than two separate
# vectors of numbers and colors) so a wavenumber and its color can never end
# up mismatched if someone later adds, removes, or reorders a peak -- each
# row is a single (position, color) pair that travels together.
# Used only by the precipitate plot (annotation_type = "peaks" below).
suberin_peak_positions <- data.frame(
  wavenumber = c(2923, 2853, # aliphatic C-H stretch
                  1739, 1712, # carbonyl C=O stretch
                  1635, 1603, # olefinics, ester-linked aromatics
                  1463, 1374, # methylene bending
                  1244, 1176, 1161, 722, # fingerprint
                  1600, 1510, 1030, 1050),  # contaminant peaks, lignin and sugars
  color = c("#e6ded6", "#e6ded6", # aliphatic C-H stretch
                 "#d1d9ec", "#d1d9ec", # carbonyl C=O stretch
                 "#dee8e0", "#dee8e0", # olefinics, ester-linked aromatics
                 "#eaeee6", "#eaeee6", # methylene bending
                 "#ece3ec", "#ece3ec", "#ece3ec", "#ece3ec", # fingerprint
                 "#ff4343", "#ff4343", "#ff4343", "#ff4343"), # contaminant peaks, lignin and sugars
  stringsAsFactors = FALSE
)

# Builds an "average + range" stacked overlay plot for an arbitrary set of plants.
#
# data:            dpt_data (or similar), filtered internally to `plants`
# plants:          character vector of plant names, bottom to top in the final plot
# methods_include: if supplied, ONLY these methods are kept (a short
#                  "allow-list" -- used for the precipitate plot, where it's
#                  easier to name the 3 methods you want than to exclude
#                  everything else). Takes priority over methods_exclude.
# methods_exclude: methods to drop before plotting (a "block-list" -- used
#                  for the main solubility-series plot). Ignored whenever
#                  methods_include is supplied, so a call never has to
#                  satisfy both rules at once.
# annotation_type: "regions" (default) shades the 5 functional-group
#                  integration windows, exactly as before. "peaks" instead
#                  draws dashed vertical lines at the suberin peak positions
#                  above -- use this for the precipitate plot.
build_stacked_range_plot <- function(data, plants, plant_palette,
                                      methods_include = NULL,
                                      methods_exclude = c("h2o20", "h2o30", "naoh20", "naoh30",
                                                           "raw", "suberin-precipitate",
                                                           "suberin-dark-precipitate"),
                                      annotation_type = c("regions", "peaks"),
                                      offset_amount = 2.8) {

  # match.arg() checks annotation_type against the two allowed values and
  # fills in "regions" if the argument is left out entirely. If you typo
  # something like "peak" (missing the s), this fails loudly right away
  # instead of quietly falling through to an `else` branch you didn't mean.
  annotation_type <- match.arg(annotation_type)

  plot_data <- data %>%
    filter(plant %in% plants)

  # Apply exactly ONE of the two filtering rules -- an include-list (keep
  # only these) or an exclude-list (drop these) -- never both, so there's
  # always one unambiguous set of methods per call.
  if (!is.null(methods_include)) {
    plot_data <- plot_data %>% filter(method %in% methods_include)
  } else {
    plot_data <- plot_data %>% filter(!method %in% methods_exclude)
  }

  plot_data <- plot_data %>%
    mutate(plant = factor(plant, levels = plants)) %>%
    mutate(snv_corrected = as.numeric(snv_corrected))

  # Calculate vertical offset for each plant to create stacking effect
  plant_offsets <- setNames(seq(0, length(plants) - 1) * offset_amount, plants)

  # Add offset to absorbance values based on plant
  plot_data <- plot_data %>%
    mutate(offset_absorbance = snv_corrected + plant_offsets[as.character(plant)])

  # Determine max y position for annotations
  max_y <- max(plot_data$offset_absorbance, na.rm = TRUE)

  # Create the base plot -- axes, theme, legend layout only. Functional-group
  # or peak annotations are added separately below, once we know which type
  # was asked for.
  #
  # CLEANUP NOTE: the original had TWO scale_x_reverse() calls stacked in the
  # same "+" chain (one with breaks starting at 4000, one at 3500). In
  # ggplot2, adding the same scale twice doesn't error and doesn't merge them
  # either -- the second one simply replaces the first. So the first call
  # here was dead code the whole time; only the second (kept below) was ever
  # actually controlling the axis. Removed the redundant one.
  plot_canvas <- ggplot() +
    labs(x = expression("Wavenumber (cm"^-1*")"),
         y = "Normalized absorbance") +
    theme(
      axis.text.y = element_blank(),
      axis.title.y = element_text(size = 15, face = "bold"),
      axis.text.x = element_text(size = 15, face = "bold"),
      axis.title.x = element_text(size = 16, face = "bold"),
      legend.position = "right",
      legend.justification = c(0, 0.01),  # (x, y) position; 0,0 is bottom-right
      legend.box = "vertical",
      legend.spacing.y = unit(1.8, "cm"),
      legend.title = element_text(size = 15, face = "bold"),
      legend.text = element_text(size = 14.5),
      plot.margin = margin(7, 14, 7, 7, "pt"),  # (top, right, bottom, left)
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank()
    ) +
    scale_y_continuous(
      breaks = plant_offsets + 0.003, # y-tick positions
      labels = NULL,
      expand = expansion(mult = c(0.003, 0.015)) # (bottom padding, top padding)
    ) +
    scale_x_reverse(
      limits = c(3750, 600),
      breaks = seq(3600, 600, -200),  # Ticks every 200 from 3600 down to 600
      expand = c(0.002, 0.003)  # Remove padding on x-axis
    )

  # Add ONE of the two annotation styles, depending on what was requested.
  if (annotation_type == "regions") {
    # Shaded rectangles marking the 5 functional-group integration windows --
    # identical to the original behavior. Used for the solubility-series plot.
    plot_canvas <- plot_canvas +
      # Simple plant / aliphatic region (two sub-windows)
      annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf,
               alpha = 0.12, fill = "#e3c8b1") +
      annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf,
               alpha = 0.12, fill = "#e3c8b1") +
      # Carbonyl region
      annotate("rect", xmin = 1780, xmax = 1690, ymin = -Inf, ymax = Inf,
               alpha = 0.18, fill = "#83a0e8") +
      # Protein/amide region
      annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf,
               alpha = 0.16, fill = "#a8cbad") +
      # Complex plant region
      annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf,
               alpha = 0.12, fill = "#beb5d5") +
      # Sugars region
      annotate("rect", xmin = 1200, xmax = 900, ymin = -Inf, ymax = Inf,
               alpha = 0.1, fill = "#c9aac7")

  } else {
    # "peaks": dashed vertical lines at known suberin-associated wavenumbers
    # (suberin_peak_positions, defined above the function). Used for the
    # precipitate plot, where the 5 functional-group windows aren't the point.
    #
    # We loop and call geom_vline() once per row with a literal `color =`
    # value (e.g. color = "#e3c8b1"), rather than mapping color with
    # aes(color = ...). This matters: aes(color = ...) would register a new
    # ggplot color SCALE, which would collide with the per-plant legend
    # scales built later in this function via ggnewscale (ggnewscale lets
    # multiple color scales coexist in one plot, but only for scales it's
    # told to isolate). A literal color argument just paints the line and
    # never touches any color scale or legend, so it can't interfere.
    for (i in seq_len(nrow(suberin_peak_positions))) {
      plot_canvas <- plot_canvas +
        geom_vline(xintercept = suberin_peak_positions$wavenumber[i],
                   color = suberin_peak_positions$color[i],
                   linetype = "solid",
                   linewidth = 0.39,
                   alpha = 1)
    }
  }

  # Add layers for each plant with average and range
  combined_plot <- plot_canvas

  for (current_plant in plants) {
    plant_data <- plot_data %>% filter(plant == current_plant)

    for (mt in unique(plant_data$method)) {
      mt_data <- plant_data %>% filter(method == mt)

      # Calculate average and range for this plant+method combination
      summary_data <- mt_data %>%
        group_by(wavenumber) %>%
        summarise(
          mean_absorbance = mean(offset_absorbance, na.rm = TRUE),
          min_absorbance = min(offset_absorbance, na.rm = TRUE),
          max_absorbance = max(offset_absorbance, na.rm = TRUE),
          .groups = "drop"
        )

      # Add range ribbon (semi-transparent)
      combined_plot <- combined_plot +
        geom_ribbon(data = summary_data,
                    aes(x = wavenumber, ymin = min_absorbance, ymax = max_absorbance),
                    fill = plant_palette[paste(current_plant, mt, sep = "_")],
                    alpha = 0.58,
                    color = NA,
                    show.legend = FALSE)

      # Add average line (opaque)
      combined_plot <- combined_plot +
        geom_line(data = summary_data,
                  aes(x = wavenumber, y = mean_absorbance),
                  color = plant_palette[paste(current_plant, mt, sep = "_")],
                  linewidth = 0.88,
                  alpha = 1,
                  show.legend = FALSE)
    }
  }

  # Build one legend per plant using invisible dummy layers + ggnewscale, so each
  # plant keeps its own color palette in its own legend block.
  # plants goes bottom-to-top in the actual plot (first element = bottom offset),
  # so the legend blocks are built in the REVERSE order to match: the plant at the top of
  # the plot gets the top legend block, working down to the plant at the bottom.
  if (!require(ggnewscale, quietly = TRUE)) {
    install.packages("ggnewscale")
    library(ggnewscale)
  }

  legend_order_plants <- rev(plants)

  for (i in seq_along(legend_order_plants)) {
    current_plant <- legend_order_plants[i]

    # Only include methods that actually have data for this plant
    plant_plot_data <- plot_data %>%
    filter(plant == current_plant)
    present_mts <- sort(unique(plant_plot_data$method))
    if (length(present_mts) == 0) next

    dummy_data <- data.frame(
      x = rep(3777, length(present_mts)),
      y = rep(0, length(present_mts)),
      mt = present_mts
    )

    combined_plot <- combined_plot +
      geom_line(data = dummy_data, aes(x = x, y = y, color = mt), linewidth = 0, alpha = 0) +
      scale_color_manual(
        name = str_to_title(current_plant),
        values =     setNames(
        plant_palette[paste(current_plant,
                            present_mts,
                            sep="_")],
        present_mts
    ),
        breaks = present_mts,
        labels = present_mts,
        guide = guide_legend(override.aes = list(linewidth = 3, alpha = 1), order = i)
      )

    if (i < length(legend_order_plants)) {
      combined_plot <- combined_plot + ggnewscale::new_scale_color()
    }
  }

  return(list(plot = combined_plot, data = plot_data))
}
```

### Plot solubility series (raw-milled, insoluble, soluble fractions)
Raw, insoluble, and all soluble fractions -- the suberin extract fractions (`suberin-precipitate`, `suberin-dark-precipitate`) are excluded here (that's now the `build_stacked_range_plot()` default) and get their own plot in the next section instead, since they're a different material class -- precipitated extract chemistry, not root residue -- and read better compared directly against `raw-milled` than folded into this series.
```{r stacked-range-plot-solubility-series}
stacked_plot <- build_stacked_range_plot(
  data = dpt_data,
  plants = plants,
  plant_palette = plant_palette,
  annotation_type = "regions"  # shaded functional-group windows (the default)
  # methods_exclude is left at its default: drops h2o/naoh trial washes,
  # unmilled "raw", and both suberin precipitate fractions.
  )

print(stacked_plot$plot)
ggsave_themed("combined_overlay_range_solubility_series.png", stacked_plot$plot, width = 13, height = 9, dpi = 300)
```

### Plot suberin extract fractions (precipitates vs. raw-milled), faceted by plant
Only `raw-milled` (as a reference spectrum) plus the two precipitate fractions. Now that just 3 methods are being compared, each plant gets its own panel (`facet_wrap`) instead of being stacked with vertical offsets on one shared axis -- so color only needs to distinguish method, not plant *and* method at once. Color ramps hue across the 3 methods at one fixed, light lightness (`get_precipitate_palette()` in `ITQB_color_config.R`, Section 7) rather than lightness within a single hue, so every line/ribbon reads as a bright, distinct color instead of a dark/mid/light shade of one color family. Dashed lines mark known suberin-associated peaks, same as before.
```{r precipitate-facet-plot}
# One row per plant x method x wavenumber, averaged across replicate DPT
# files -- same mean/min/max summary shape used by the material-class plot
# above, just scoped to the 3 precipitate-figure methods instead of all
# plot_group values.
precip_only_methods <- setdiff(precipitate_methods, "raw-milled")

plants_with_precipitates <- dpt_data %>%
  filter(method %in% precip_only_methods) %>%
  distinct(plant) %>%
  pull(plant)

precipitate_plot_data <- dpt_data %>%
  filter(method %in% precipitate_methods, plant %in% plants_with_precipitates) %>%
  group_by(plant, method, wavenumber) %>%
  summarise(
    mean_absorbance = mean(snv_corrected, na.rm = TRUE),
    min_absorbance  = min(snv_corrected, na.rm = TRUE),
    max_absorbance  = max(snv_corrected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Factor with an explicit level order (rather than default alphabetical)
  # so the legend and any faceting/grouping always lists methods the same
  # way, and so it lines up with precipitate_palette's names.
  mutate(method = factor(method, levels = precipitate_methods))

precipitate_palette <- get_precipitate_palette()

precipitate_facet_plot <- ggplot(precipitate_plot_data,
                                  aes(x = wavenumber, color = method, fill = method)) +
  geom_ribbon(aes(ymin = min_absorbance, ymax = max_absorbance), alpha = 0.3, color = NA) +
  geom_line(aes(y = mean_absorbance), linewidth = 0.76) +
  scale_x_reverse(limits = c(3750, 600), breaks = seq(3600, 600, -200), expand = c(0.002, 0.003)) +
  scale_color_manual(values = precipitate_palette, name = "Method") +
  scale_fill_manual(values = precipitate_palette, name = "Method") +
  facet_wrap(~ str_to_title(plant), ncol = 1) +
  labs(x = expression("Wavenumber (cm"^-1*")"), y = "Normalized absorbance") +
  theme(strip.text = element_text(face = "bold", size = 12),
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        plot.margin = margin(7, 14, 7, 7, "pt"),  # (top, right, bottom, left)
  )

# Dashed suberin-peak markers, added the SAME way as in build_stacked_range_plot()
# (a for-loop of geom_vline() calls, re-assigning the plot each time) -- each
# call uses a literal `color =`, not aes(color = ...), so these lines stay
# outside the method color/fill scales above instead of adding extra legend
# entries or fighting with scale_color_manual()/scale_fill_manual().
for (i in seq_len(nrow(suberin_peak_positions))) {
  precipitate_facet_plot <- precipitate_facet_plot +
    geom_vline(xintercept = suberin_peak_positions$wavenumber[i],
               color = suberin_peak_positions$color[i],
               linetype = "solid", linewidth = 0.3, alpha = 0.98)
}

print(precipitate_facet_plot)
ggsave_themed("precipitate_facet_by_plant.png", precipitate_facet_plot, dpi = 300)
```

### Ester : aromatic peak ratio (~1740 cm<sup>-1</sup> : ~1510 cm<sup>-1</sup>)
The ~1740 cm<sup>-1</sup> band sits right where the suberin ester-carbonyl peak is marked above (1739 cm<sup>-1</sup> in `suberin_peak_positions`); ~1510 cm<sup>-1</sup> falls inside the aromatic/lignin window used elsewhere in this document (1550-1500 cm<sup>-1</sup>, "Complex Plant"). Rather than reading a single point (sensitive to instrument noise at exactly that wavenumber), `find_peak_height()` below takes the LOCAL MAXIMUM absorbance within a small window around each target, per sample -- closer to how you'd pick a peak by eye. This uses SNV-normalized `SNV_corrected`, matching how `perform_spectral_integration()` computes the 5 functional-group proportions elsewhere in this document, rather than `snv_corrected` (which is tuned for visual overlay, not for preserving relative peak heights).
```{r peak-ratio-function}
#' Find the local-maximum absorbance near a target wavenumber, per sample.
#' @param data          data frame with columns wavenumber, absorbance, filename (e.g. dpt_data)
#' @param target        center wavenumber (cm^-1) to search around
#' @param window         +/- cm^-1 tolerance searched for the local max
#'                       (ASSUMPTION: 8 cm^-1 is a starting guess -- widen/narrow
#'                       once you've checked it lands on the actual peak apex
#'                       for your spectra, not a shoulder or noise spike)
#' @param value_col     which column holds the intensity to maximize (default: SNV-normalized absorbance)
#' @return one row per filename: filename, wavenumber_at_peak, peak_absorbance
find_peak_height <- function(data, target, window = 8, value_col = "snv_corrected") {
  data %>%
    filter(wavenumber >= target - window, wavenumber <= target + window) %>%
    group_by(filename) %>%
    slice_max(order_by = .data[[value_col]], n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(filename,
              wavenumber_at_peak = wavenumber,
              peak_absorbance = .data[[value_col]])
}
```

```{r peak-ratio-calculation}
# Restrict to the same 3 methods shown in the faceted plot above -- widen
# this filter later if you want the ratio for other methods too, nothing
# else in this chunk assumes only 3 methods.
precip_dpt_data <- dpt_data %>% filter(method %in% precipitate_methods, plant %in% plants_with_precipitates)

peak_1740 <- find_peak_height(precip_dpt_data, target = 1740) %>%
  dplyr::rename(abs_1740 = peak_absorbance, wavenumber_1740 = wavenumber_at_peak)
peak_1510 <- find_peak_height(precip_dpt_data, target = 1510) %>%
  dplyr::rename(abs_1510 = peak_absorbance, wavenumber_1510 = wavenumber_at_peak)

# One row per sample: plant/method metadata + both peak heights + their ratio.
sample_metadata <- precip_dpt_data %>%
  select(filename, plant, method) %>%
  distinct()

peak_ratio_data <- sample_metadata %>%
  left_join(peak_1740, by = "filename") %>%
  left_join(peak_1510, by = "filename") %>%
  mutate(ester_aromatic_ratio = abs_1740 / abs_1510)

# Mean +/- SE per plant x method, same shape as plant_method_stats earlier.
peak_ratio_stats <- peak_ratio_data %>%
  group_by(plant, method) %>%
  summarise(
    mean_ratio = mean(ester_aromatic_ratio, na.rm = TRUE),
    SE_ratio   = sd(ester_aromatic_ratio, na.rm = TRUE) / sqrt(n()),
    n          = n(),
    .groups = "drop"
  ) %>%
  mutate(method = factor(method, levels = precipitate_methods),
         plant  = str_to_title(plant)) %>%
  arrange(plant, method)

knitr::kable(peak_ratio_stats, digits = 3,
             col.names = c("Plant", "Method", "Mean ratio (1740:1510)", "SE", "n"))
```

```{r peak-ratio-plot, fig.width=8, fig.height=6}
peak_ratio_plot <- ggplot(peak_ratio_stats, aes(x = plant, y = mean_ratio, fill = method)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean_ratio - SE_ratio, ymax = mean_ratio + SE_ratio),
                position = position_dodge(width = 0.8), width = 0.2, color = "white") +
  scale_fill_manual(values = precipitate_palette, name = "Method") +
  labs(x = NULL, y = "Ester (~1740) : Aromatic (~1510) ratio",
       title = "Ester-to-aromatic peak ratio") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()
  )

print(peak_ratio_plot)
ggsave_themed("precipitate_peak_ratio.png", peak_ratio_plot, width = 8, height = 6, dpi = 300)
```



## Plotting by material class
Raw and insoluble each get their own flat hue; all soluble fractions share one hue but get shaded lighter→darker by round, reusing the same region_lightness_ranks() trick your functional-group bars already use (so 1-2 vs. 3-4 vs. 5+ read as a visible progression, not three arbitrary colors). A mean ± range overlay per plant, colored by plot_group, is the cleanest way to show raw → insoluble → progressively-depleted soluble fractions on one axis.

```{r material-class-plot}
plot_data <- dpt_data %>%
  filter(!is.na(plot_group)) %>%   # drops "other" (h2o/naoh) -- see note in Section 5
  group_by(plant, plot_group, wavenumber) %>%
  summarise(
    mean_absorbance = mean(snv_corrected, na.rm = TRUE),
    min_absorbance  = min(snv_corrected, na.rm = TRUE),
    max_absorbance  = max(snv_corrected, na.rm = TRUE),
    .groups = "drop"
  )

material_series_plot <- ggplot(plot_data, aes(x = wavenumber, color = plot_group, fill = plot_group)) +
  geom_ribbon(aes(ymin = min_absorbance, ymax = max_absorbance), alpha = 0.3, color = NA) +
  geom_line(aes(y = mean_absorbance), linewidth = 0.9) +
  scale_x_reverse(limits = c(3750, 600), breaks = seq(3600, 600, -200), expand = c(0.001, 0.001)) +
  scale_color_manual(values = material_bin_palette, name = "Fraction") +
  scale_fill_manual(values = material_bin_palette, name = "Fraction") +
  facet_wrap(~ str_to_title(plant), ncol = 1) +
  labs(x = expression("Wavenumber (cm"^-1*")"), y = "Normalized absorbance") +
  theme(strip.text = element_text(face = "bold", size = 12),
  axis.text.y = element_blank()
  )

print(material_series_plot)
ggsave_themed("material_class_extraction_series.png", material_series_plot, width = 9, height = 11, dpi = 300)
```

# Integration of functional regions
Modified from "Berhe_Ghezzehei_Lab/DRIFTS Data Analysis/Visualizing_FTIR_w_Elemental_Data.R" which starts by calculating aliphatic and aromatic indices, then deriving simple and complex plant and protein/amide proportions before visualizing everything. This procedure sums non-normalized absorbance values over a specified range, and divides by the total peak area in the ranges of interest to get the proportions.

## Performing the integration
```{r spectral-integration-function}
# Function to perform spectral integration on DPT data (based on Leila_code_modified.Rmd)
perform_spectral_integration <- function(dpt_data) {
  cat("Starting spectral integration processing...\n")
  
  # Load required library for integration
  if (!require(pracma, quietly = TRUE)) {
    install.packages("pracma")
    library(pracma)
  }
  
  # Prepare the data in the format needed for integration
  # Rename columns to match expected format
  comb <- dpt_data %>%
    select(filename, wavenumber, snv_corrected) %>%
    dplyr::rename(sample = filename, abs = snv_corrected)
  
  # Define spectral windows of interest (same as Leila's code)
  red1 <- comb[comb$wavenumber > 2839 & comb$wavenumber < 2870, ]  # Aliphatic 1
  red2 <- comb[comb$wavenumber > 2898 & comb$wavenumber < 2976, ]  # Aliphatic 2  
  red3 <- comb[comb$wavenumber > 1690 & comb$wavenumber < 1780, ]  # Carbonyl
  red4 <- comb[comb$wavenumber > 1580 & comb$wavenumber < 1660, ]  # Protein/amide
  red5 <- comb[comb$wavenumber > 1500 & comb$wavenumber < 1550, ]  # Complex plant
  red6 <- comb[comb$wavenumber > 900 & comb$wavenumber < 1200, ]   # Sugars
  
  # Get unique samples
  sample <- unique(comb$sample)
  
  # Create integration results data frame
  ints <- data.frame(
    sample = sample,
    int_2839_2870 = numeric(length(sample)),
    int_2898_2976 = numeric(length(sample)),
    int_1690_1780 = numeric(length(sample)),
    int_1580_1660 = numeric(length(sample)),
    int_1500_1550 = numeric(length(sample)),
    int_900_1200 = numeric(length(sample)),
    stringsAsFactors = FALSE
  )
  
  # Calculate integrated areas for each spectral window and sample
  cat("Processing", length(sample), "samples for integration...\n")
  for (i in seq_len(nrow(ints))) {
    current_sample <- sample[i]
    
    # Extract data for current sample from each spectral window
    sample_red1 <- red1[red1$sample == current_sample, ]
    sample_red2 <- red2[red2$sample == current_sample, ]
    sample_red3 <- red3[red3$sample == current_sample, ]
    sample_red4 <- red4[red4$sample == current_sample, ]
    sample_red5 <- red5[red5$sample == current_sample, ]
    sample_red6 <- red6[red6$sample == current_sample, ]
    
    # Calculate area under curve (integration) using trapezoidal rule
    if (nrow(sample_red1) > 1) {
      ints$int_2839_2870[i] <- pracma::trapz(sample_red1$wavenumber, sample_red1$abs)
    }
    
    if (nrow(sample_red2) > 1) {
      ints$int_2898_2976[i] <- pracma::trapz(sample_red2$wavenumber, sample_red2$abs)
    }
    
    if (nrow(sample_red3) > 1) {
      ints$int_1690_1780[i] <- pracma::trapz(sample_red3$wavenumber, sample_red3$abs)
    }
    
    if (nrow(sample_red4) > 1) {
      ints$int_1580_1660[i] <- pracma::trapz(sample_red4$wavenumber, sample_red4$abs)
    }
    
    if (nrow(sample_red5) > 1) {
      ints$int_1500_1550[i] <- pracma::trapz(sample_red5$wavenumber, sample_red5$abs)
    }
    
    if (nrow(sample_red6) > 1) {
      ints$int_900_1200[i] <- pracma::trapz(sample_red6$wavenumber, sample_red6$abs)
    }
  }
  
  # Use existing metadata from the input data (no need to re-extract)
  cat("Using existing metadata from DPT data...\n")
  basic_meta <- dpt_data %>%
    select(filename, plant, method, run_date, run_rep) %>%
    distinct() %>%
    dplyr::rename(sample = filename)
  
  # Combine metadata with integration results
  ftir_data <- merge(basic_meta, ints, by = "sample", all = TRUE)
  
  # Rename integration columns to match expected names
  names(ftir_data)[names(ftir_data) == "int_1580_1660"] <- "int_microbial"
  names(ftir_data)[names(ftir_data) == "int_1500_1550"] <- "int_complex_plant"
  names(ftir_data)[names(ftir_data) == "int_1690_1780"] <- "int_carbonyl"
  names(ftir_data)[names(ftir_data) == "int_900_1200"] <- "int_sugars"
  
  # Calculate derived metrics
  ftir_data$aliphatic <- ftir_data$int_2839_2870 + ftir_data$int_2898_2976
  ftir_data$total_peak_area <- ftir_data$aliphatic + ftir_data$int_microbial + ftir_data$int_complex_plant + ftir_data$int_carbonyl + ftir_data$int_sugars
  ftir_data$aliphatic_prop <- ftir_data$aliphatic / ftir_data$total_peak_area
  ftir_data$complex_plant_prop <- ftir_data$int_complex_plant / ftir_data$total_peak_area
  ftir_data$carbonyl_prop <- ftir_data$int_carbonyl / ftir_data$total_peak_area
  ftir_data$sugars_prop <- ftir_data$int_sugars / ftir_data$total_peak_area
  ftir_data$microbial_prop <- ftir_data$int_microbial / ftir_data$total_peak_area
  ftir_data$aliphatic_microbial_prop <- ftir_data$aliphatic_prop / ftir_data$microbial_prop
  ftir_data$complex_microbial_prop <- ftir_data$complex_plant_prop / ftir_data$microbial_prop
  ftir_data$total_area_no_sugars <- ftir_data$total_peak_area - ftir_data$int_sugars
  ftir_data$aliphatic_prop_no_sugars <- ftir_data$aliphatic / ftir_data$total_area_no_sugars
  ftir_data$complex_plant_prop_no_sugars <- ftir_data$int_complex_plant / ftir_data$total_area_no_sugars
  # Remove individual aliphatic integration columns (keep only derived metrics)
  ftir_data <- ftir_data %>%
    select(-int_2839_2870, -int_2898_2976)
  
  # Save the processed data to the output directory
  today <- format(Sys.Date(), "%Y-%m-%d")
  output_file <- file.path(outputs_folder, paste0("processed_int_data_", today, ".csv"))
  write.csv(ftir_data, file = output_file, row.names = FALSE)
  cat("Saved processed integration data to:", output_file, "\n")
  
  cat("Integration processing completed successfully.\n")
  return(ftir_data)
}
# Try to find the most recent processed integration data file
ftir_file <- find_most_recent_processed_file(outputs_folder)

use_cache <- FALSE

if (!is.null(ftir_file) && file.exists(ftir_file)) {
  cached_ftir_data <- read_csv(ftir_file, show_col_types = FALSE)

  # ftir_data$sample is just dpt_data$filename renamed (see basic_meta above),
  # so we can directly compare the DPT files just loaded against the files
  # already represented in the cached integration results.
  current_files <- unique(dpt_data$filename)
  cached_files  <- unique(cached_ftir_data$sample)
  new_files     <- setdiff(current_files, cached_files)

  if (length(new_files) == 0) {
    ftir_data <- cached_ftir_data
    use_cache <- TRUE
    cat("No new DPT files since cache was generated -- reusing cached integration data from:", basename(ftir_file), "\n")
    cat("Dimensions:", dim(ftir_data), "\n")
    cat("Columns:", names(ftir_data), "\n")
  } else {
    cat("Found", length(new_files), "new DPT file(s) not in the cached data:\n")
    cat(paste(" -", new_files), sep = "\n")
    cat("\nReprocessing all DPT data to include the new file(s)...\n")
  }
} else {
  cat("No processed integration data found. Processing raw DPT data to generate integration results...\n")
}

if (!use_cache) {
  # Check if we have the required dpt_data from previous chunk
  if (!exists("dpt_data") || nrow(dpt_data) == 0) {
    stop("No DPT data available for integration. Please ensure the DPT data import was successful.")
  }
  
  # Process raw DPT data using the integrated function
  ftir_data <- perform_spectral_integration(dpt_data)
  
  # Verify the processing was successful
  if (is.null(ftir_data) || nrow(ftir_data) == 0) {
    stop("Failed to process DPT data for integration.")
  }
  
  cat("Successfully processed", nrow(ftir_data), "samples from DPT data\n")
  cat("Generated columns:", names(ftir_data), "\n")
}
```

## Generating statistics
This code chunk calculates summary statistics (mean and standard error) for the derived metrics (aliphatic, carbonyl, protein/amide, complex plant, sugars proportions) grouped by plant, preparation method, and plant-method combinations. The results are stored in separate data frames for later visualization.

```{r integration-statistics}
# Calculate summary statistics by group
plant_stats <- ftir_data %>%
  mutate(plant = factor(plant, levels = sort(unique(plant)))) %>%
  group_by(plant) %>%
  summarise(
    Mean_aliphatic_prop = mean(aliphatic_prop, na.rm = TRUE),
    SE_aliphatic_prop = sd(aliphatic_prop, na.rm = TRUE) / sqrt(n()),
    Mean_carbonyl_prop = mean(carbonyl_prop, na.rm = TRUE),
    SE_carbonyl_prop = sd(carbonyl_prop, na.rm = TRUE) / sqrt(n()),
    Mean_microbial_prop = mean(microbial_prop, na.rm = TRUE),
    SE_microbial_prop = sd(microbial_prop, na.rm = TRUE) / sqrt(n()),
    Mean_complex_plant_prop = mean(complex_plant_prop, na.rm = TRUE),
    SE_complex_plant_prop = sd(complex_plant_prop, na.rm = TRUE) / sqrt(n()),
    Mean_sugars_prop = mean(sugars_prop, na.rm = TRUE),
    SE_sugars_prop = sd(sugars_prop, na.rm = TRUE) / sqrt(n()),
    Mean_aliphatic_prop_no_sugars = mean(aliphatic_prop_no_sugars, na.rm = TRUE),
    SE_aliphatic_prop_no_sugars = sd(aliphatic_prop_no_sugars, na.rm = TRUE) / sqrt(n()),
    Mean_carbonyl_prop_no_sugars = mean(carbonyl_prop, na.rm = TRUE),
    SE_carbonyl_prop_no_sugars = sd(carbonyl_prop, na.rm = TRUE) / sqrt(n()),
    Mean_microbial_prop_no_sugars = mean(microbial_prop, na.rm = TRUE),
    SE_microbial_prop_no_sugars = sd(microbial_prop, na.rm = TRUE) / sqrt(n()),
    Mean_complex_plant_prop_no_sugars = mean(complex_plant_prop_no_sugars, na.rm = TRUE),
    SE_complex_plant_prop_no_sugars = sd(complex_plant_prop_no_sugars, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

method_stats <- ftir_data %>%
  group_by(method) %>%
  summarise(
    Mean_aliphatic_prop = mean(aliphatic_prop, na.rm = TRUE),
    SE_aliphatic_prop = sd(aliphatic_prop, na.rm = TRUE) / sqrt(n()),
    Mean_carbonyl_prop = mean(carbonyl_prop, na.rm = TRUE),
    SE_carbonyl_prop = sd(carbonyl_prop, na.rm = TRUE) / sqrt(n()),
    Mean_microbial_prop = mean(microbial_prop, na.rm = TRUE),
    SE_microbial_prop = sd(microbial_prop, na.rm = TRUE) / sqrt(n()),
    Mean_complex_plant_prop = mean(complex_plant_prop, na.rm = TRUE),
    SE_complex_plant_prop = sd(complex_plant_prop, na.rm = TRUE) / sqrt(n()),
    Mean_sugars_prop = mean(sugars_prop, na.rm = TRUE),
    SE_sugars_prop = sd(sugars_prop, na.rm = TRUE) / sqrt(n()),
    Mean_aliphatic_prop_no_sugars = mean(aliphatic_prop_no_sugars, na.rm = TRUE),
    SE_aliphatic_prop_no_sugars = sd(aliphatic_prop_no_sugars, na.rm = TRUE) / sqrt(n()),
    Mean_carbonyl_prop_no_sugars = mean(carbonyl_prop, na.rm = TRUE),
    SE_carbonyl_prop_no_sugars = sd(carbonyl_prop, na.rm = TRUE) / sqrt(n()),
    Mean_microbial_prop_no_sugars = mean(microbial_prop, na.rm = TRUE),
    SE_microbial_prop_no_sugars = sd(microbial_prop, na.rm = TRUE) / sqrt(n()),
    Mean_complex_plant_prop_no_sugars = mean(complex_plant_prop_no_sugars, na.rm = TRUE),
    SE_complex_plant_prop_no_sugars = sd(complex_plant_prop_no_sugars, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

plant_method_stats <- ftir_data %>%
  mutate(plant = factor(plant, levels = sort(unique(plant)))) %>%
  group_by(plant, method) %>%
  summarise(
    Mean_aliphatic_prop = mean(aliphatic_prop, na.rm = TRUE),
    SE_aliphatic_prop = sd(aliphatic_prop, na.rm = TRUE) / sqrt(n()),
    Mean_carbonyl_prop = mean(carbonyl_prop, na.rm = TRUE),
    SE_carbonyl_prop = sd(carbonyl_prop, na.rm = TRUE) / sqrt(n()),
    Mean_microbial_prop = mean(microbial_prop, na.rm = TRUE),
    SE_microbial_prop = sd(microbial_prop, na.rm = TRUE) / sqrt(n()),
    Mean_complex_plant_prop = mean(complex_plant_prop, na.rm = TRUE),
    SE_complex_plant_prop = sd(complex_plant_prop, na.rm = TRUE) / sqrt(n()),
    Mean_sugars_prop = mean(sugars_prop, na.rm = TRUE),
    SE_sugars_prop = sd(sugars_prop, na.rm = TRUE) / sqrt(n()),
    Mean_aliphatic_prop_no_sugars = mean(aliphatic_prop_no_sugars, na.rm = TRUE),
    SE_aliphatic_prop_no_sugars = sd(aliphatic_prop_no_sugars, na.rm = TRUE) / sqrt(n()),
    Mean_carbonyl_prop_no_sugars = mean(carbonyl_prop, na.rm = TRUE),
    SE_carbonyl_prop_no_sugars = sd(carbonyl_prop, na.rm = TRUE) / sqrt(n()),
    Mean_microbial_prop_no_sugars = mean(microbial_prop, na.rm = TRUE),
    SE_microbial_prop_no_sugars = sd(microbial_prop, na.rm = TRUE) / sqrt(n()),
    Mean_complex_plant_prop_no_sugars = mean(complex_plant_prop_no_sugars, na.rm = TRUE),
    SE_complex_plant_prop_no_sugars = sd(complex_plant_prop_no_sugars, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )
```

# Visualizing functional group proportions by method
Percentage stacked bar charts comparing preparation methods within each plant. Each bar is one method; segments are the five functional-group region proportions (aliphatic/simple plant, carbonyl, protein/amide, complex plant, sugars) from `plant_method_stats`. Colors come from `get_method_region_palette()`: each method keeps its own fixed hue (from `method_colors`/`method_hues` in the config), and region is encoded as a lightness shade within that hue, ordered so adjacent segments in the stack are never two similar shades.

```{r method-region-bar-function}
# Region display/stacking order -- bottom to top. This order also drives the
# color assignment in get_method_region_palette(), so change it here only.
region_order <- c("aliphatic", "carbonyl", "microbial", "complex", "sugar")
region_labels <- c(
  aliphatic = "Simple Plant",
  carbonyl  = "Carbonyl",
  microbial = "Protein/amide",
  complex   = "Complex Plant",
  sugar     = "Sugars"
)

# Lightness assigned to each region -- the SAME values used to build the fill
# colors (get_method_region_palette()) and the grayscale key (build_region_key()).
# Reused below to pick label text color, so labels always match the key's logic
# even if region_order or method_region_L_range change.
region_lightness <- get_region_lightness(region_order)


# Methods to leave out of the functional-group bar charts entirely: the
# h2o/naoh trial washes, unmilled "raw", and both suberin precipitate
# fractions. The precipitates are extract chemistry rather than root
# residue, so integrating them under these 5 root-derived windows wouldn't
# be a meaningful comparison.
# Defined ONCE here and reused below in the combined (all-plant) figure too --
# previously the combined figure computed its method list straight from
# plant_method_stats with no exclusion filter at all, so a method dropped
# from every individual per-plant chart could still have leaked back into
# the combined one. Sharing this single vector between both makes that
# impossible: whichever methods change here update both figures identically.
excluded_bar_methods <- c("h2o20", "h2o30", "naoh20", "naoh30", "raw",
                           "suberin-precipitate", "suberin-dark-precipitate")

# Builds one plant's % stacked bar chart, one bar per method.
build_method_region_bar_plot <- function(stats_data, current_plant, title = NULL) {
  plant_data <- stats_data %>%
    filter(plant == current_plant) %>%
    filter(!method %in% excluded_bar_methods)


  present_methods <- sort(unique(plant_data$method))

  if (is.null(title)) {
    title <- str_to_title(current_plant)
  }

  long_data <- plant_data %>%
    select(method,
           aliphatic = Mean_aliphatic_prop,
           carbonyl  = Mean_carbonyl_prop,
           microbial = Mean_microbial_prop,
           complex   = Mean_complex_plant_prop,
           sugar     = Mean_sugars_prop) %>%
    pivot_longer(cols = -method, names_to = "region", values_to = "prop") %>%
    mutate(
      region   = factor(region, levels = region_order),
      method   = factor(method, levels = present_methods),
      pct      = prop * 100,
      fill_key = paste(method, region, sep = "_"),
      text_color = ifelse(region_lightness[as.character(region)] > text_color_L_threshold, "black", "white")
    )

  region_palette <- get_method_region_palette(regions = region_order, methods = present_methods)

  ggplot(long_data, aes(x = method, y = prop, fill = fill_key, group = region)) +
    geom_col(position = position_stack(reverse = TRUE),
             linewidth = 0.00, width = 0.88) +
    geom_text(aes(label = paste0(round(pct, 1), "%")),
              position = position_stack(reverse = TRUE, vjust = 0.5),
              color = long_data$text_color,
              size = 3.6, fontface = "bold") +
    scale_fill_manual(values = region_palette) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0.01, 0.02))) +
    guides(fill = "none") +
    labs(x = NULL, y = "Functional Group Proportion (%)", title = title) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
          legend.position = "none")
}
```

Because region is encoded only as lightness (not hue), one grayscale key explains the shading for every method/plant panel -- there's no need for a separate legend per panel.

```{r method-region-bar-plots, fig.width=4, fig.height=6}
plants_for_bar <- sort(unique(plant_method_stats$plant))

region_key_plot <- build_region_key(region_order, region_labels)

bar_plots <- lapply(plants_for_bar, function(pt) {
  build_method_region_bar_plot(plant_method_stats, pt)
})
names(bar_plots) <- plants_for_bar

for (pt in plants_for_bar) {
  pt_plot_with_key <- cowplot::plot_grid(
    bar_plots[[pt]], region_key_plot,
    nrow = 1,
    align = "h", axis = "tb",  # lines up the key's y-axis with the bar plot's y-axis
    rel_widths = c(1, 0.3)    # same idea as the combined figure's rel_widths, just for 1 plant instead of N
  )

  print(pt_plot_with_key)
  ggsave_themed(paste0("method_region_stackedbar_", pt, ".png"),
                pt_plot_with_key, width = 5.5, height = 6, dpi = 300)
}
```

### Combined figure (all plants side-by-side)
Unlike the individual per-plant plots above (each a separate ggplot, combined via `cowplot::plot_grid`), this version builds ONE ggplot across all plants and uses `facet_grid(~ plant, scales = "free_x", space = "free_x")` so every panel shares a single y-axis, panel widths scale with each plant's actual method count (so panels sit close together instead of getting equal width regardless of bar count), and the plant name (the facet strip) is centered over its panel automatically. Only methods present for **every** plant are shown, so each panel is directly comparable method-for-method; any method missing from even one plant is dropped and reported in the console output.
```{r method-region-bar-combined, fig.width=13, fig.height=6}
# Only include methods present in EVERY selected plant (intersect, not union) --
# a method missing from even one plant is dropped from the combined figure
# entirely, so panels are directly comparable method-for-method.
#
# excluded_bar_methods (h2o/naoh trials, raw, both suberin precipitates) is
# applied FIRST, before the intersect. Previously this chunk built its method
# list straight from plant_method_stats with no exclusion at all, so a
# method dropped from every individual per-plant bar chart above could still
# reappear here if it happened to be present for every plant -- e.g. the
# precipitate fractions, which exist for every plant, would have snuck back
# in. Filtering with the same shared list the per-plant plots use closes
# that gap.
methods_by_plant <- lapply(plants_for_bar, function(pt) {
  pt_methods <- plant_method_stats$method[plant_method_stats$plant == pt]
  setdiff(unique(pt_methods), excluded_bar_methods)
})
present_methods_all <- sort(Reduce(intersect, methods_by_plant))

# Split what got dropped into two groups for the console message, so
# deliberately-excluded methods (expected, no action needed) aren't reported
# the same way as methods that are genuinely just missing for one plant
# (worth noticing, might be an incomplete dataset).
all_methods_seen <- sort(unique(plant_method_stats$method[plant_method_stats$plant %in% plants_for_bar]))
dropped_methods  <- setdiff(all_methods_seen, present_methods_all)
dropped_missing  <- setdiff(dropped_methods, excluded_bar_methods)
if (length(dropped_missing) > 0) {
  cat("Methods not present for every plant (excluded from combined figure):",
      paste(dropped_missing, collapse = ", "), "\n")
}

combined_region_palette <- get_method_region_palette(regions = region_order, methods = present_methods_all)

combined_long_data <- plant_method_stats %>%
  filter(plant %in% plants_for_bar, method %in% present_methods_all) %>%
  mutate(plant = factor(str_to_title(as.character(plant)),
                        levels = str_to_title(plants_for_bar))) %>%
  select(plant, method,
         aliphatic = Mean_aliphatic_prop,
         carbonyl  = Mean_carbonyl_prop,
         microbial = Mean_microbial_prop,
         complex   = Mean_complex_plant_prop,
         sugar     = Mean_sugars_prop) %>%
  pivot_longer(cols = -c(plant, method), names_to = "region", values_to = "prop") %>%
  mutate(
    region     = factor(region, levels = region_order),
    method     = factor(method, levels = present_methods_all),
    pct        = prop * 100,
    fill_key   = paste(method, region, sep = "_"),
    text_color = ifelse(region_lightness[as.character(region)] > text_color_L_threshold,
                         "black", "white")
  )

combined_bars <- ggplot(combined_long_data, aes(x = method, y = prop, fill = fill_key, group = region)) +
  geom_col(position = position_stack(reverse = TRUE), linewidth = 0.00, width = 0.88) +
  geom_text(aes(label = paste0(round(pct, 1), "%"), color = text_color),
            position = position_stack(reverse = TRUE, vjust = 0.5),
            size = 3.6, fontface = "bold") +
  scale_fill_manual(values = combined_region_palette) +
  scale_color_identity() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0.01, 0.02))) +
  facet_grid(~ plant, scales = "free_x", space = "free_x") +
  guides(fill = "none") +
  labs(x = NULL, y = "Functional Group Proportion (%)") +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    panel.spacing   = unit(4, "pt"),
    strip.text      = element_text(face = "bold", size = 12, hjust = 0.5)
  )

combined_bar_plot <- cowplot::plot_grid(
  combined_bars, region_key_plot,
  nrow = 1,
  align = "h", axis = "tb", # make the key's y-axis line up with the bars' y-axis
  rel_widths = c(length(plants_for_bar), 0.4)
)

print(combined_bar_plot)
ggsave_themed("method_region_stackedbar_all_plants.png", combined_bar_plot,
              width = 13, height = 6, dpi = 300)
```

### Version of the combined figure without sugars
This version of the combined figure excludes the sugars region from the stacked bars, so the remaining four regions (aliphatic/simple plant, carbonyl, protein/amide, complex plant) are scaled to sum to 100% for each method. This allows for a more direct comparison of the non-sugar functional groups across methods and plants.
```{r method-region-bar-combined-no-sugars, fig.width=13, fig.height=6}
