---
title: "ATR Analysis - ITQB"
author: "Adrianne Seiden"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: darkly
    highlight: zenburn
    code_folding: hide
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding,
                     output_dir = "../adrianne-code/") })
---
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  results = "markup",
  include = TRUE
)
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))
## Installing packages ##
list_of_packages <- c(
  "plyr",
  "dplyr",
  "stringr",
  "reshape",
  "data.table",
  "nc",
  "readr",
  "tidyr",
  "pracma"
)

# install missing packages
new_packages <- list_of_packages[!(
  list_of_packages %in% installed.packages()[, "Package"]
)]
if (length(new_packages)) install.packages(new_packages)

# load packages
library(plyr)
library(dplyr)
library(stringr)
library(reshape)
library(data.table)
library(nc)
library(readr)
library(tidyr)
library(pracma)

# Set paths
dpt_folder <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/all_abs_DPT_files"
integrated_ftir_csv <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/ATR_outputs_csv"
integrated_ftir_plots <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/ATR_outputs_plots"
```
```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))


# Set paths

integrated_ftir_data <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/ATR_outputs_csv"
outputs_folder <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/ATR_outputs_plots"

# Create output directory if it doesn't exist
if (!dir.exists(outputs_folder)) {
  dir.create(outputs_folder, recursive = TRUE)
}

# Helper function to find the most recent processed integration data file
find_most_recent_processed_file <- function(data_dir) {
  pattern <- "processed_int_data_.*\\.csv$"
  files <- list.files(data_dir, pattern = pattern, full.names = TRUE)
  
  if (length(files) == 0) {
    return(NULL)
  }
  
  # Extract dates from filenames and find the most recent
  dates <- stringr::str_extract(basename(files), "\\d{4}-\\d{2}-\\d{2}")
  most_recent_idx <- which.max(as.Date(dates))
  
  cat("Found", length(files), "processed integration files\n")
  cat("Using most recent file:", basename(files[most_recent_idx]), "\n")
  
  return(files[most_recent_idx])
}

# Load libraries
library(ggplot2)
library(ggridges)
library(readr)
library(stringr)
library(viridis)
library(dplyr)
library(cowplot)
library(grid)
library(ggsignif)
library(tidyr)

# Install and load required packages for integration if needed
required_packages <- c("pracma", "nc")
for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    cat("Installing package:", pkg, "\n")
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}

# Source centralized theme and color configuration
source(file.path(dirname(knitr::current_input()), "config/ITQB_theme_colors_config.R"))

# Ensure theme is set and variables are initialized
# theme_set(theme_dark_custom())
timepoint_colors <- c("#50478A", "#3D6988", "#3D8770", "#75873D", "#876B3D")
type_colors <- c("root" = "#8B4513", "soil" = "#DEB887")
```

# Overview

This document integrates analysis of ATR DPT files from OPUSprocessing.Rmd and Leila_code_modified.Rmd.

1. **OPUSprocessing.Rmd "Reading DPT files" section** - Ridge plots showing spectral data by crop with different washing/extraction treatments.
2. **Leila_code_modified.Rmd** - Bar plots showing functional group proportions 

# Preparing data for analysis

## Importing DPT data
This code is modified from code by Stephany Soledad Chacon, "Box> Salk Institute Project> Salk Data> FTIR_Intact_decomposition_pots> FTIR_analysis.Rmd". Using data in the DPT format, we can extrapolate sample attributes from the file names, then normalize and visualize spectra across replicates, crops, and treatments. The width of the line bounding the spectra represents the range of the data.
```{r import-dpt-data}
# Check if folder exists and list files
if (!dir.exists(dpt_folder)) {
  stop("Directory does not exist: ", dpt_folder)
}

# List and sort files
cat("Files found in directory:\n")
print(list.files(dpt_folder))
DPTfiles <- sort(list.files(dpt_folder, pattern = ".dpt"))
cat("Number of .dpt files:", length(DPTfiles), "\n")

# Read files individually and combine
dpt_data_list <- list()

for (i in seq_along(DPTfiles)) {
  file_path <- file.path(dpt_folder, DPTfiles[i])
  
  # Read as lines and parse manually (DPT files are space-delimited)
  lines <- readLines(file_path, warn = FALSE)
  lines <- lines[lines != ""]  # Remove empty lines
  
  # Split each line by whitespace
  split_lines <- strsplit(lines, "\\s+")
  
  # Convert to data frame
  temp_data <- tryCatch({
    data.frame(
      wavenumber = as.numeric(sapply(split_lines, `[`, 1)),
      absorbance = as.numeric(sapply(split_lines, `[`, 2)),
      stringsAsFactors = FALSE
    )
  }, error = function(e) NULL)
  
  if (!is.null(temp_data)) {
    # Add filename column
    temp_data$filename <- DPTfiles[i]
    
    # Store in list
    dpt_data_list[[i]] <- temp_data
  }
}

# Remove NULL entries (failed reads)
dpt_data_list <- dpt_data_list[!sapply(dpt_data_list, is.null)]

# Combine all data
dpt_data <- bind_rows(dpt_data_list)

# Remove any rows with NA values
dpt_data <- dpt_data[complete.cases(dpt_data), ]

# Extract metadata from filenames

# Helper to extract fields from filenames with the pattern
# AKS_suberin_[plant]_[washing-treatment]_[yymmdd]-[run].dpt
extract_filename_fields <- function(filename) {
  stem <- tools::file_path_sans_ext(basename(filename))
  matches <- stringr::str_match(
    stem,
    "^AKS_suberin_([^_]+)_([^_]+)_([0-9]{6})-([0-9]+)$"
  )

  if (is.na(matches[1, 1])) {
    return(tibble(
      plant = NA_character_,
      washing_treatment = NA_character_,
      run_date = NA_character_,
      run_rep = NA_integer_
    ))
  }

  tibble(
    plant = tolower(matches[1, 2]),
    washing_treatment = tolower(matches[1, 3]),
    run_date = as.Date(matches[1, 4], format = "%y%m%d"),
    run_rep = as.integer(matches[1, 5])
  )
}

file_metadata <- bind_rows(lapply(dpt_data$filename, extract_filename_fields))
dpt_data <- bind_cols(dpt_data, file_metadata)

# Preserve the later normalization code without changing its object name.
dpt_data_roots <- dpt_data

# save as csv for record-keeping
write.csv(dpt_data, file.path(outputs_folder, "dpt_data.csv"), row.names = FALSE)


cat("Samples by plant:\n")
print(table(dpt_data_roots$plant))
cat("Samples by treatment:\n") 
print(table(dpt_data_roots$washing_treatment))
```

## Normalizing data
This code is modified from code by Stephany Soledad Chacon, "Box> Salk Institute Project> Salk Data> FTIR_Intact_decomposition_pots> FTIR_analysis.Rmd". It creates a new column with normalized absorbance values for each sample, using baseline correction (subtracting the minimum) followed by normalization to the maximum absorbance value.
```{r normalize-data}
# Normalize absorbance data by sample (same approach as OPUSprocessing.Rmd)
dpt_data_roots <- dpt_data_roots %>%
  group_by(filename) %>%
  mutate(
    # Baseline correction - subtract minimum
    baseline_corrected = absorbance - min(absorbance, na.rm = TRUE),
    # Normalize to maximum
    normalized_absorbance = baseline_corrected / max(baseline_corrected, na.rm = TRUE)
  ) %>%
  ungroup()
```


# Plotting ATR spectra
## Plant-stacked overlay with treatment shading
The plot below mirrors the combined stacked overlay from the DRIFTS analysis, but groups by plant and shades by washing treatment. Treatment order is explicit so the lightest lines are raw and the darkest are NaOH30.

```{r combined-stacked-overlay-no-noplant-range, fig.height=9, fig.width=13}
treatment_levels <- c("raw", "h2o20", "h2o30", "naoh20", "naoh30")
treatment_colors <- c(
  "raw" = "#e6e6e6",
  "h2o20" = "#c7dcef",
  "h2o30" = "#8fb7d8",
  "naoh20" = "#4f86b5",
  "naoh30" = "#1f4e79"
)

plant_levels <- sort(unique(dpt_data_roots$plant))
atr_plot_data <- dpt_data_roots %>%
  mutate(
    plant = factor(plant, levels = plant_levels),
    washing_treatment = factor(washing_treatment, levels = treatment_levels),
    normalized_absorbance = as.numeric(normalized_absorbance)
  ) %>%
  filter(!is.na(plant), !is.na(washing_treatment))

offset_amount_plant <- 0.8
plant_offsets <- setNames(seq(0, (length(plant_levels) - 1) * offset_amount_plant, offset_amount_plant), plant_levels)

atr_plot_data <- atr_plot_data %>%
  mutate(offset_absorbance = normalized_absorbance + plant_offsets[as.character(plant)])

max_y_atr <- max(atr_plot_data$offset_absorbance, na.rm = TRUE)

base_plot_atr <- ggplot() +
  labs(
    x = expression("Wavenumber (cm"^-1*")"),
    y = "Normalized absorbance"
  ) +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_text(size = 15, face = "bold"),
    axis.text.x = element_text(size = 15, face = "bold"),
    axis.title.x = element_text(size = 16, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 13.5),
    plot.margin = margin(7, 14, 7, 7, "pt"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  scale_y_continuous(
    breaks = plant_offsets + 0.003,
    labels = names(plant_offsets),
    expand = expansion(mult = c(0.003, 0.02))
  ) +
  scale_x_reverse(
    limits = c(3770, 415),
    breaks = seq(3500, 500, -500),
    expand = c(0.002, 0.003)
  ) +
  geom_vline(xintercept = 1500, linetype = "dotted", linewidth = 0.46, color = annotation_text_color, alpha = 0.8) +
  geom_vline(xintercept = 1550, linetype = "dotted", linewidth = 0.46, color = annotation_text_color, alpha = 0.8) +
  geom_vline(xintercept = 1580, linetype = "dashed", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
  geom_vline(xintercept = 1660, linetype = "dashed", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
  geom_vline(xintercept = 2839, linetype = "longdash", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
  geom_vline(xintercept = 2870, linetype = "longdash", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
  geom_vline(xintercept = 2898, linetype = "longdash", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
  geom_vline(xintercept = 2976, linetype = "longdash", linewidth = 0.23, color = annotation_text_color, alpha = 0.8) +
  annotate("rect", xmin = 2976, xmax = 2898, ymin = -Inf, ymax = Inf, alpha = 0.12, fill = "#e3c8b1") +
  annotate("rect", xmin = 2870, xmax = 2839, ymin = -Inf, ymax = Inf, alpha = 0.12, fill = "#e3c8b1") +
  annotate("text", x = 2560, y = max_y_atr, label = "Simple Plant", size = 3.4, face = "bold", hjust = 0.5, color = annotation_text_color) +
  annotate("rect", xmin = 1660, xmax = 1580, ymin = -Inf, ymax = Inf, alpha = 0.12, fill = "#a8cbad") +
  annotate("text", x = 1870, y = max_y_atr + 0.08, label = "Microbial", size = 3.4, face = "bold", hjust = 0.5, color = annotation_text_color) +
  annotate("rect", xmin = 1550, xmax = 1500, ymin = -Inf, ymax = Inf, alpha = 0.12, fill = "#beb5d5") +
  annotate("text", x = 1190, y = max_y_atr + 0.08, label = "Complex Plant", size = 3.4, face = "bold", hjust = 0.5, color = annotation_text_color)

combined_overlay_range_plot <- base_plot_atr

for (plant_name in plant_levels) {
  plant_data <- atr_plot_data %>% filter(plant == plant_name)

  for (treatment_name in treatment_levels) {
    treatment_data <- plant_data %>% filter(washing_treatment == treatment_name)
    if (nrow(treatment_data) == 0) next

    summary_data <- treatment_data %>%
      group_by(wavenumber) %>%
      summarise(
        mean_absorbance = mean(offset_absorbance, na.rm = TRUE),
        min_absorbance = min(offset_absorbance, na.rm = TRUE),
        max_absorbance = max(offset_absorbance, na.rm = TRUE),
        .groups = "drop"
      )

    combined_overlay_range_plot <- combined_overlay_range_plot +
      geom_ribbon(
        data = summary_data,
        aes(x = wavenumber, ymin = min_absorbance, ymax = max_absorbance),
        fill = treatment_colors[[treatment_name]],
        alpha = 0.18,
        inherit.aes = FALSE
      ) +
      geom_line(
        data = summary_data,
        aes(x = wavenumber, y = mean_absorbance, color = treatment_name),
        linewidth = 0.7,
        inherit.aes = FALSE
      )
  }
}

combined_overlay_range_plot <- combined_overlay_range_plot +
  scale_color_manual(
    name = "Washing treatment",
    values = treatment_colors,
    breaks = treatment_levels,
    labels = c("Raw", "H2O20", "H2O30", "NaOH20", "NaOH30"),
    guide = guide_legend(override.aes = list(linewidth = 2, alpha = 1))
  )

print(combined_overlay_range_plot)
```


## Spectral integration
The integration step is set up so you can change the windows in one place before recomputing the areas. Update the endpoints below if you want to use different functional-group ranges for ATR.

```{r spectral-integration, message=FALSE, warning=FALSE}
# Edit these ranges if you want different ATR integration windows.
integration_windows <- list(
  simple_aliphatic_1 = c(2976, 2898),
  simple_aliphatic_2 = c(2870, 2839),
  microbial = c(1660, 1580),
  complex_plant = c(1550, 1500)
)

integration_samples <- dpt_data_roots %>%
  filter(!is.na(plant), !is.na(washing_treatment)) %>%
  distinct(filename, plant, washing_treatment, run_date, run_rep)

integration_results <- integration_samples

for (window_name in names(integration_windows)) {
  window_limits <- integration_windows[[window_name]]
  window_data <- dpt_data_roots %>%
    filter(wavenumber < max(window_limits), wavenumber > min(window_limits))

  window_values <- window_data %>%
    group_by(filename) %>%
    summarise(
      integrated_area = pracma::trapz(wavenumber, absorbance),
      .groups = "drop"
    )

  names(window_values)[2] <- paste0("int_", window_name)
  integration_results <- left_join(integration_results, window_values, by = "filename")
}

integration_results <- integration_results %>%
  mutate(
    simple_aliphatic = int_simple_aliphatic_1 + int_simple_aliphatic_2,
    total_peak_area = simple_aliphatic + int_microbial + int_complex_plant,
    simple_plant_prop = simple_aliphatic / total_peak_area,
    complex_plant_prop = int_complex_plant / total_peak_area,
    microbial_prop = int_microbial / total_peak_area
  )

write.csv(integration_results, file.path(outputs_folder, "atr_integrated_peak_data.csv"), row.names = FALSE)
```