---
title: "Jar Flux Analysis"
author: "Adrianne Seiden"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: darkly
    highlight: zenburn
    code_folding: hide
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding,
                     output_dir = "../adrianne-code/") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))

# Set Box path
box_base <- Sys.getenv("BOX_BASE")
if (box_base == "") stop("BOX_BASE environment variable is not set!")

# Set paths
csv_folder <- file.path(box_base, "Salk Institute Project/AKS Salk files/CO2 data")
outputs_folder <- file.path(box_base, "Salk Institute Project/AKS Salk files/CO2 data/confounding-factors-outputs")

# Load required libraries
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(purrr)
library(ggplot2)

# Source centralized theme and color configuration
source(file.path(dirname(knitr::current_input()), "config/theme_colors_config.R"))

# Ensure theme is set and variables are initialized
theme_set(theme_dark_custom())
```

## Load Data

```{r load-data}
# Load both sheets from the Excel file
file_path <- file.path(csv_folder, "jar_flux_moisture_csv.xlsx")

# Load CO2 data
jar_co2 <- read_excel(file_path, sheet = "Jars") %>%
  mutate(
    sampling_date = as_date(sampling_date),
    jar_num = as.character(jar_num)
  )

# Load moisture data
jar_moisture <- read_excel(file_path, sheet = "moisture-data") %>%
  mutate(
    watering_date = as_date(watering_date),
    jar_num = as.character(jar_num)
  )

# Display data structure
glimpse(jar_co2)
glimpse(jar_moisture)

# Check the date ranges and number of observations
cat("\n=== DATA DIAGNOSTICS ===\n")
cat("Number of rows in jar_co2:", nrow(jar_co2), "\n")
cat("Number of unique jars in jar_co2:", length(unique(jar_co2$jar_num)), "\n")
cat("Number of unique sampling dates in jar_co2:", length(unique(jar_co2$sampling_date)), "\n")
cat("\nUnique CO2 sampling dates:\n")
print(sort(unique(jar_co2$sampling_date)))

cat("\nNumber of rows in jar_moisture:", nrow(jar_moisture), "\n")
cat("Number of unique jars in jar_moisture:", length(unique(jar_moisture$jar_num)), "\n")
cat("Number of unique watering dates in jar_moisture:", length(unique(jar_moisture$watering_date)), "\n")
cat("\nUnique watering dates:\n")
print(sort(unique(jar_moisture$watering_date)))

# Check for any date formatting issues
cat("\nSample of sampling_date values (first 10):\n")
print(head(jar_co2$sampling_date, 10))
```

## Calculate GWC for Each CO2 Sampling Date

The approach:
1. For CO2 sampling dates that match a watering date, use the pre-watering GWC
2. For dates between watering events, interpolate between the post-watering water content (previous watering) and pre-watering water content (next watering)

### Interpolation Methods

Multiple interpolation methods available:

- **Linear**: Simple straight-line interpolation
  - *Assumes*: Constant drying rate
  - *Best for*: Short time intervals or conservative estimates
  
- **Exponential** (RECOMMENDED): Exponential decay model
  - *Assumes*: Drying rate proportional to current moisture (most realistic for soil)
  - *Best for*: Natural soil drying processes
  - *Formula*: Water follows exponential decay from post-watering to pre-watering values
  
- **Logarithmic**: Log-space interpolation
  - *Assumes*: Multiplicative changes in water content
  - *Best for*: When percentage changes are more consistent than absolute changes

**Recommendation**: Use **exponential** for most soil applications, as it best captures the physics of evaporative drying where the rate decreases as soil becomes drier.

```{r calculate-gwc}
# Choose interpolation method: "linear", "exponential", or "logarithmic"
interpolation_method <- "exponential"

# Function to calculate GWC for a single jar on a single date
calculate_gwc_for_jar <- function(jar_id, sample_date, moisture_data, method) {
  # Get moisture data for this specific jar
  jar_moisture_data <- moisture_data %>% 
    filter(jar_num == jar_id) %>%
    arrange(watering_date)
  
  # Check if no moisture data exists for this jar
  if (nrow(jar_moisture_data) == 0) {
    return(NA_real_)
  }
  
  # Check if sampling date exactly matches a watering date (use pre-watering GWC)
  exact_match <- jar_moisture_data %>%
    filter(watering_date == sample_date) %>%
    pull(`GWC_pre-watering_g/g`)
  
  if (length(exact_match) > 0) {
    # Exact match found - use pre-watering GWC
    return(exact_match[1])
  }
  
  # Interpolate between watering events
  # Find the most recent watering date before sampling_date
  prev_watering <- jar_moisture_data %>%
    filter(watering_date < sample_date) %>%
    arrange(desc(watering_date)) %>%
    slice(1)
  
  # Find the next watering date after sampling_date
  next_watering <- jar_moisture_data %>%
    filter(watering_date > sample_date) %>%
    arrange(watering_date) %>%
    slice(1)
  
  if (nrow(prev_watering) == 0 && nrow(next_watering) == 0) {
    # No watering data available
    return(NA_real_)
  } else if (nrow(prev_watering) == 0) {
    # Before first watering - use first pre-watering GWC
    return(next_watering$`GWC_pre-watering_g/g`[1])
  } else if (nrow(next_watering) == 0) {
    # After last watering - use last post-watering GWC
    return(prev_watering$`GWC_post-watering`[1])
  }
  
  # Interpolate between prev post-watering and next pre-watering
  # Get the water content values
  water_prev_post <- prev_watering$`water_post-watering_g`[1]
  water_next_pre <- next_watering$`water_pre-watering_g`[1]
  
  # Get the dates
  date_prev <- prev_watering$watering_date[1]
  date_next <- next_watering$watering_date[1]
  
  # Get dry soil mass (should be constant for a jar)
  dry_soil_mass <- prev_watering$`Mass_dry_soil_g_no-root`[1]
  
  # Calculate time fraction (0 to 1)
  time_fraction <- as.numeric(sample_date - date_prev) / 
                  as.numeric(date_next - date_prev)
  
  # Apply selected interpolation method
  water_interpolated <- if (method == "linear") {
    # Linear interpolation (constant drying rate)
    water_prev_post + time_fraction * (water_next_pre - water_prev_post)
    
  } else if (method == "exponential") {
    # Exponential decay (realistic for soil drying)
    if (water_prev_post <= water_next_pre) {
      # Water is increasing or staying same (shouldn't happen, but use linear)
      water_prev_post + time_fraction * (water_next_pre - water_prev_post)
    } else {
      # Normal case: water is decreasing
      W_min <- 0.01  # Minimum water content (nearly dry)
      
      if (water_next_pre <= W_min) {
        # If next value is at minimum, use linear
        water_prev_post + time_fraction * (water_next_pre - water_prev_post)
      } else {
        # Calculate decay constant
        k <- -log((water_next_pre - W_min) / (water_prev_post - W_min))
        # Interpolate using exponential decay
        W_min + (water_prev_post - W_min) * exp(-k * time_fraction)
      }
    }
    
  } else if (method == "logarithmic") {
    # Logarithmic interpolation
    if (water_prev_post <= 0 || water_next_pre <= 0) {
      # Can't take log of negative/zero, use linear
      water_prev_post + time_fraction * (water_next_pre - water_prev_post)
    } else {
      # Interpolate in log space
      log_start <- log(water_prev_post)
      log_end <- log(water_next_pre)
      exp(log_start + time_fraction * (log_end - log_start))
    }
    
  } else {
    # Default to linear if method not recognized
    water_prev_post + time_fraction * (water_next_pre - water_prev_post)
  }
  
  # Convert to GWC
  gwc_interpolated <- water_interpolated / dry_soil_mass
  
  return(gwc_interpolated)
}

# For each jar and CO2 sampling date, estimate GWC
gwc_estimates <- jar_co2 %>%
  select(jar_num, sampling_date) %>%
  distinct() %>%
  mutate(
    estimated_GWC = purrr::map2_dbl(
      jar_num, 
      sampling_date, 
      ~calculate_gwc_for_jar(.x, .y, jar_moisture, interpolation_method)
    )
  )

# Display the estimates
cat("\nFirst 20 GWC estimates (showing jar-specific values):\n")
head(gwc_estimates, 20) %>% print()

# Check for variety in estimates
cat("\nGWC estimates by jar for first sampling date:\n")
gwc_estimates %>%
  filter(sampling_date == min(sampling_date)) %>%
  arrange(jar_num) %>%
  head(10) %>%
  print()

# Summary statistics
cat("\nInterpolation method used:", interpolation_method, "\n")
cat("Summary of estimated GWC:\n")
print(summary(gwc_estimates$estimated_GWC))
cat("\nNumber of unique GWC values:", length(unique(gwc_estimates$estimated_GWC)), "\n")
```

## Join GWC Back to CO2 Data

```{r join-gwc}
# Check gwc_estimates before join
cat("\n=== BEFORE JOIN ===\n")
cat("Rows in gwc_estimates:", nrow(gwc_estimates), "\n")
cat("Unique dates in gwc_estimates:", length(unique(gwc_estimates$sampling_date)), "\n")
cat("Date range in gwc_estimates:", as.character(range(gwc_estimates$sampling_date)), "\n")

# Join the estimated GWC to the original jar_co2 data
jar_co2_with_gwc <- jar_co2 %>%
  left_join(gwc_estimates, by = c("jar_num", "sampling_date")) %>%
  # Replace the empty GWC column with our estimates
  mutate(GWC = estimated_GWC) %>%
  select(-estimated_GWC)

# Check the results
cat("\n=== GWC RESULTS DIAGNOSTICS ===\n")
cat("Number of rows in jar_co2_with_gwc:", nrow(jar_co2_with_gwc), "\n")
cat("Number of unique dates in jar_co2_with_gwc:", length(unique(jar_co2_with_gwc$sampling_date)), "\n")
cat("Date class:", class(jar_co2_with_gwc$sampling_date), "\n")
cat("Unique dates in final data:\n")
print(sort(unique(jar_co2_with_gwc$sampling_date)))

cat("\nGWC summary:\n")
summary(jar_co2_with_gwc$GWC)

cat("\nNumber of non-NA GWC values:", sum(!is.na(jar_co2_with_gwc$GWC)), "\n")
cat("Number of NA GWC values:", sum(is.na(jar_co2_with_gwc$GWC)), "\n")

# View data across multiple dates
cat("\nExample data from multiple dates:\n")
jar_co2_with_gwc %>%
  filter(jar_num %in% c("1", "2")) %>%
  select(jar_num, sampling_date, crop, sampling_wk, GWC) %>%
  arrange(jar_num, sampling_date) %>%
  print(n = 25)
```

## Visualize GWC Over Time

```{r plot-gwc, fig.width=12, fig.height=6}
library(ggplot2)

# Check data before plotting
cat("\n=== PLOT DATA CHECK ===\n")
cat("Total rows in jar_co2_with_gwc:", nrow(jar_co2_with_gwc), "\n")
cat("Date class:", class(jar_co2_with_gwc$sampling_date), "\n")
cat("Date range:", as.character(min(jar_co2_with_gwc$sampling_date, na.rm = TRUE)), 
    "to", as.character(max(jar_co2_with_gwc$sampling_date, na.rm = TRUE)), "\n")
cat("Unique dates:", length(unique(jar_co2_with_gwc$sampling_date)), "\n")

# Find jars with the most sampling dates (to show trends)
jar_sampling_counts <- jar_co2_with_gwc %>%
  group_by(jar_num) %>%
  summarise(
    n_dates = n_distinct(sampling_date),
    first_date = min(sampling_date),
    last_date = max(sampling_date),
    .groups = "drop"
  ) %>%
  arrange(desc(n_dates))

cat("Jars with most sampling dates:\n")
print(head(jar_sampling_counts, 30))

# Select jars that were sampled until the last date (kept until the end)
last_sampling_date <- max(jar_co2_with_gwc$sampling_date)
jars_kept_to_end <- jar_co2_with_gwc %>%
  filter(sampling_date == last_sampling_date) %>%
  pull(jar_num) %>%
  unique()

cat("\nNumber of jars sampled on last date (", as.character(last_sampling_date), "):", 
    length(jars_kept_to_end), "\n")
cat("Jar numbers:", paste(sort(jars_kept_to_end), collapse=", "), "\n")

# Use all 24 jars that were kept to the end
example_jars <- jars_kept_to_end

plot_data <- jar_co2_with_gwc %>%
  filter(jar_num %in% example_jars)

cat("Rows in plot_data:", nrow(plot_data), "\n") 
cat("Dates in plot_data:", length(unique(plot_data$sampling_date)), "\n")
cat("Date range in plot_data:", as.character(range(plot_data$sampling_date)), "\n")

# Show a sample of the plot data
cat("\nSample of plot_data:\n")
plot_data %>%
  select(jar_num, sampling_date, GWC) %>%
  arrange(jar_num, sampling_date) %>%
  print(n = 20)

# Create plot - with 24 jars, facet by crop for clarity
ggplot(plot_data, aes(x = sampling_date, y = GWC, color = crop, group = jar_num)) +
  geom_line(linewidth = 0.5, alpha = 0.7) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_manual(values = crop_colors) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  labs(
    title = paste("Gravimetric Water Content Over Time - All Jars Kept to End (n=", 
                  length(example_jars), ", ", interpolation_method, " interpolation)", sep=""),
    x = "Sampling Date",
    y = "GWC (g water / g dry soil)",
    color = "Crop Type"
  ) +
  theme_dark() +
  theme(
    plot.background = element_rect(fill = "#0B0B0B", color = NA),
    panel.background = element_rect(fill = "#111111", color = NA),
    panel.grid.major = element_line(color = "#3A3A3A", linewidth = 0.25),
    panel.grid.minor = element_line(color = "#242424", linewidth = 0.2),
    axis.text = element_text(color = "#F5F5F5"),
    axis.title = element_text(color = "#FFFFFF"),
    plot.title = element_text(color = "#FFFFFF", face = "bold"),
    legend.background = element_rect(fill = "#0B0B0B", color = NA),
    legend.key = element_rect(fill = "#111111", color = NA),
    legend.text = element_text(color = "#F5F5F5"),
    legend.title = element_text(color = "#FFFFFF"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Average GWC by Crop Over Time (24 Jars Kept to End)

```{r plot-gwc-crop-average, fig.width=12, fig.height=6}
# Average GWC across jars for each crop and sampling date
avg_gwc_by_crop <- plot_data %>%
  group_by(crop, sampling_date) %>%
  summarise(
    mean_GWC = mean(GWC, na.rm = TRUE),
    sd_GWC = sd(GWC, na.rm = TRUE),
    n_jars = sum(!is.na(GWC)),
    se_GWC = if_else(n_jars > 1, sd_GWC / sqrt(n_jars), NA_real_),
    .groups = "drop"
  )

cat("\nRows in avg_gwc_by_crop:", nrow(avg_gwc_by_crop), "\n")
cat("Crops in avg_gwc_by_crop:", paste(unique(avg_gwc_by_crop$crop), collapse = ", "), "\n")

ggplot(avg_gwc_by_crop, aes(x = sampling_date, y = mean_GWC, color = crop, group = crop)) +
  geom_errorbar(
    aes(ymin = mean_GWC - se_GWC, ymax = mean_GWC + se_GWC),
    width = 2,
    linewidth = 0.5,
    alpha = 0.8
  ) +
  geom_line(linewidth = 1.1, alpha = 0.95) +
  geom_point(size = 2.3, alpha = 0.95) +
  scale_color_manual(values = crop_colors) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  labs(
    title = "Average Gravimetric Water Content Over Time by Crop (24 End-Kept Jars)",
    x = "Sampling Date",
    y = "Average GWC (g water / g dry soil)",
    color = "Crop Type"
  ) +
  theme_dark() +
  theme(
    plot.background = element_rect(fill = "#0B0B0B", color = NA),
    panel.background = element_rect(fill = "#111111", color = NA),
    panel.grid.major = element_line(color = "#3A3A3A", linewidth = 0.25),
    panel.grid.minor = element_line(color = "#242424", linewidth = 0.2),
    axis.text = element_text(color = "#F5F5F5"),
    axis.title = element_text(color = "#FFFFFF"),
    plot.title = element_text(color = "#FFFFFF", face = "bold"),
    legend.background = element_rect(fill = "#0B0B0B", color = NA),
    legend.key = element_rect(fill = "#111111", color = NA),
    legend.text = element_text(color = "#F5F5F5"),
    legend.title = element_text(color = "#FFFFFF"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Cumulative Moving Average GWC by Crop

```{r plot-gwc-crop-moving-average, fig.width=12, fig.height=6}
# Moving average = average of all prior dates (including current date) for each crop
moving_avg_by_crop <- avg_gwc_by_crop %>%
  arrange(crop, sampling_date) %>%
  group_by(crop) %>%
  mutate(
    moving_avg_GWC = cummean(mean_GWC),
    n_points = row_number(),
    moving_sd_GWC = purrr::map_dbl(row_number(), ~sd(mean_GWC[1:.x], na.rm = TRUE)),
    moving_se_GWC = if_else(n_points > 1, moving_sd_GWC / sqrt(n_points), NA_real_)
  ) %>%
  ungroup()

ggplot(moving_avg_by_crop, aes(x = sampling_date, y = moving_avg_GWC, color = crop, group = crop)) +
  geom_errorbar(
    aes(ymin = moving_avg_GWC - moving_se_GWC, ymax = moving_avg_GWC + moving_se_GWC),
    width = 2,
    linewidth = 0.5,
    alpha = 0.8
  ) +
  geom_line(linewidth = 1.2, alpha = 0.95) +
  geom_point(size = 2.4, alpha = 0.95) +
  scale_color_manual(values = crop_colors) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  labs(
    title = "Cumulative Moving Average GWC by Crop (24 End-Kept Jars)",
    subtitle = "Each point is the mean of all crop-average values up to that date",
    x = "Sampling Date",
    y = "Cumulative moving average GWC (g water / g dry soil)",
    color = "Crop Type"
  ) +
  theme_dark() +
  theme(
    plot.background = element_rect(fill = "#0B0B0B", color = NA),
    panel.background = element_rect(fill = "#111111", color = NA),
    panel.grid.major = element_line(color = "#3A3A3A", linewidth = 0.25),
    panel.grid.minor = element_line(color = "#242424", linewidth = 0.2),
    axis.text = element_text(color = "#F5F5F5"),
    axis.title = element_text(color = "#FFFFFF"),
    plot.title = element_text(color = "#FFFFFF", face = "bold"),
    plot.subtitle = element_text(color = "#DADADA"),
    legend.background = element_rect(fill = "#0B0B0B", color = NA),
    legend.key = element_rect(fill = "#111111", color = NA),
    legend.text = element_text(color = "#F5F5F5"),
    legend.title = element_text(color = "#FFFFFF"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Compare Interpolation Methods

Let's compare how different interpolation methods affect the GWC estimates for a single jar.

```{r compare-methods, fig.width=12, fig.height=8}
# Calculate GWC with all three methods using our function
gwc_linear <- jar_co2 %>%
  select(jar_num, sampling_date) %>%
  distinct() %>%
  mutate(
    estimated_GWC = purrr::map2_dbl(
      jar_num, 
      sampling_date, 
      ~calculate_gwc_for_jar(.x, .y, jar_moisture, "linear")
    ),
    method = "linear"
  )

gwc_exponential <- jar_co2 %>%
  select(jar_num, sampling_date) %>%
  distinct() %>%
  mutate(
    estimated_GWC = purrr::map2_dbl(
      jar_num, 
      sampling_date, 
      ~calculate_gwc_for_jar(.x, .y, jar_moisture, "exponential")
    ),
    method = "exponential"
  )

gwc_logarithmic <- jar_co2 %>%
  select(jar_num, sampling_date) %>%
  distinct() %>%
  mutate(
    estimated_GWC = purrr::map2_dbl(
      jar_num, 
      sampling_date, 
      ~calculate_gwc_for_jar(.x, .y, jar_moisture, "logarithmic")
    ),
    method = "logarithmic"
  )

# Combine results
gwc_comparison <- bind_rows(gwc_linear, gwc_exponential, gwc_logarithmic)

cat("\n=== COMPARISON DATA CHECK ===\n")
cat("Total rows in gwc_comparison:", nrow(gwc_comparison), "\n")
cat("Unique dates in comparison:", length(unique(gwc_comparison$sampling_date)), "\n")
cat("Date range:", as.character(min(gwc_comparison$sampling_date, na.rm = TRUE)), 
    "to", as.character(max(gwc_comparison$sampling_date, na.rm = TRUE)), "\n")

# Use the same jars that were kept to the end
last_sampling_date <- max(gwc_comparison$sampling_date)
jars_kept_to_end <- gwc_comparison %>%
  filter(sampling_date == last_sampling_date, method == "linear") %>%  # Just one method to avoid duplication
  pull(jar_num) %>%
  unique()

# Select a subset for comparison (4-6 jars for readability in facets)
example_jars_compare <- sort(jars_kept_to_end)[1:6]

cat("Plotting jars for comparison:", paste(example_jars_compare, collapse=", "), "\n")

# Plot comparison
comparison_plot_data <- gwc_comparison %>%
  filter(jar_num %in% example_jars_compare)

cat("Rows in comparison plot:", nrow(comparison_plot_data), "\n")

comparison_plot_data %>%
  ggplot(aes(x = sampling_date, y = estimated_GWC, color = method, group = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~jar_num, scales = "free_y", ncol = 2) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b %y") +
  labs(
    title = "Comparison of Interpolation Methods Across Jars",
    subtitle = "Exponential decay typically best represents soil drying behavior",
    x = "Sampling Date",
    y = "Estimated GWC (g/g)",
    color = "Method"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Calculate differences between methods
method_differences <- gwc_comparison %>%
  select(jar_num, sampling_date, method, estimated_GWC) %>%
  pivot_wider(names_from = method, values_from = estimated_GWC) %>%
  mutate(
    diff_linear_exponential = abs(linear - exponential),
    diff_linear_logarithmic = abs(linear - logarithmic),
    diff_exponential_logarithmic = abs(exponential - logarithmic)
  )

cat("\nAverage absolute differences between methods:\n")
cat("Linear vs Exponential:", 
    mean(method_differences$diff_linear_exponential, na.rm = TRUE), "\n")
cat("Linear vs Logarithmic:", 
    mean(method_differences$diff_linear_logarithmic, na.rm = TRUE), "\n")
cat("Exponential vs Logarithmic:", 
    mean(method_differences$diff_exponential_logarithmic, na.rm = TRUE), "\n")
```

## Save Results

```{r save-results}
# Save the updated data
output_file <- file.path(csv_folder, "jar_co2_with_gwc.csv")
write.csv(jar_co2_with_gwc, output_file, row.names = FALSE)

cat("Data saved to:", output_file, "\n")
```

