---
title: "Visualizing OPUS binary files in R"
author: "Adrianne Seiden"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: darkly
    highlight: zenburn
    code_folding: hide
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding,
                     output_dir = "../adrianne-code/") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  echo = TRUE,
  results = "markup",
  include = TRUE,
  message = FALSE,
  warning = FALSE
)


# set corr_spectra path
corr_spectra <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/all_abs_OPUS_files"
# set dpt_folder path
dpt_folder <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/all_abs_DPT_files"
# set outputs folder path
outputs_folder <- "C:/Users/adria/OneDrive - University of California Merced/Summer26/Portugal/Data/FTIR_code-outputs"
```

# Reading OPUS files
[source](https://spectral-cockpit.github.io/opusreader2/articles/opusreader2_introduction.html)

This code is modified from the original OPUSreader2 documentation [vignettes/opusreader2_introduction.Rmd](https://github.com/spectral-cockpit/opusreader2/blob/main/vignettes/opusreader2_introduction.Rmd) by Philip Baumann and Thomas Knecht. It reads OPUS binary files and extracts metadata and absorbance data, which can then be plotted.

## Plots \& Metadata
The main function, `read_opus()`, reads one or more OPUS files and returns a nested list of class `list_opusreader2`. Each list contains both the spectral data and metadata for each file. The `dsn` argument is the data source name. It can be a character vector of folder paths (to read files recursively) or specific OPUS file paths. Start by testing the read_opus function on your corrected spectra folder, and saving the output of the first file.

```{r setup-first-sample, message=FALSE, warning=FALSE, echo=TRUE}
#load opusreader2 from github.com/spectral-cockpit
library(opusreader2)

#save the path to your corrected spectra
# # Mac path
# corr_spectra <- "/Users/adrianneseiden/Library/CloudStorage/Box-Box/Salk Institute Project/AKS Salk files/Adrianne_FTIRdata/Corrected_files" # nolint
#  # Windows path
# corr_spectra <- "C:/Users/adria/Box/Box-Box/Salk Institute Project/AKS Salk files/Adrianne_FTIRdata/Corrected_files" # nolint
# # Relative path
# corr_spectra <- "../Adrianne_FTIRdata/Corrected_files"

#save the data from your corrected files as 'data_test'
data_test <- read_opus(dsn = corr_spectra)
#check the names of the list
names(data_test)
# define 'meas_1' as the first element of the 'data_test' list
meas_1 <- data_test[[1]]
```

Next I defined a function, plotSpectrum, to plot the spectral data from a single sample (meas_1)

```{r fun-plot-spectrum-and-metadata, echo=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# data is a list containing OPUS file data, including absorbance and metadata.

plotSpectrum <- function(data) { # nolint: object_name_linter.
  ab_data <- data$ab
  if (!is.null(ab_data) &&
        !is.null(ab_data$wavenumbers) && !is.null(ab_data$data) &&
        is.numeric(ab_data$wavenumbers) && is.numeric(ab_data$data) &&
        length(ab_data$wavenumbers) == length(ab_data$data) &&
        all(is.finite(ab_data$wavenumbers)) && all(is.finite(ab_data$data))) {
    plot(
      ab_data$wavenumbers, ab_data$data, type = "l",
      xlab = "Wavenumber (cm⁻¹)", ylab = "Absorbance",
      main = data$basic_metadata$dsn_filename,
      xlim = rev(range(ab_data$wavenumbers))
    )
  } else {
    cat("Absorbance data not found or not valid in this file.\n")
    str(ab_data)
  }
}
plotSpectrum(meas_1)
```

I then defined the ***metadataTable*** function, to extract and display the a subset of the metadata stored in the data extracted by 'read_opus' in a table format. 

To view all the available data categories, run 'names(meas_1)' in the console. To view the parameters within these categories use str(meas_1$category_name) where 'category_name' is 'basic_metadata', 'optics', etc.
```{r fun-metadata-table, include=FALSE, echo = TRUE, message=FALSE, warning=FALSE}
## Safely extract each value (use NA if missing)
library(knitr)
getval <- function(x) if (!is.null(x) && length(x) > 0) x else NA
metadataTable <- function(data) { # nolint: object_name_linter.
  meta_vec <- c(
    `File Name` = getval(data$basic_metadata$dsn_filename),
    `Timestamp` = getval(data$basic_metadata$timestamp_string),
    `Max Y` = getval(data$ab_data_param$parameters$MXY$parameter_value),
    `Min Y` = getval(data$ab_data_param$parameters$MNY$parameter_value),
    `Aperture Setting` = getval(data$optics$parameters$APT$parameter_value),
    `Scanner Velocity` = getval(data$optics$parameters$VEL$parameter_value),
    `Result Spectrum` = getval(data$acquisition$parameters$PLF$parameter_value),
    `Resolution` = getval(data$acquisition$parameters$RES$parameter_value),
    `Sample Scans` = getval(data$acquisition$parameters$NSS$parameter_value),
    `End Frequency` =
      getval(data$fourier_transformation$parameters$HFQ$parameter_value),
    `Start Frequency` =
      getval(data$fourier_transformation$parameters$LFQ$parameter_value),
    `Duration` = getval(data$instrument$parameters$DUR$parameter_value),
    `Experiment (method)` = getval(data$sample$parameters$EXP$parameter_value)
  )
  meta_long <- data.frame(
    Parameter = names(meta_vec),
    Value = unname(meta_vec),
    row.names = NULL,
    check.names = FALSE
  )

  kable(meta_long, caption = "Metadata", align = "l")
}
metadataTable(meas_1)
```

The ***plotMetadata*** function combines the plot and table for a single spectrum.
```{r single-spectrum-metadata, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
plotMetadata <- function(data) { # nolint: object_name_linter.
  plotSpectrum(data)
  metadataTable(data)
}
plotMetadata(meas_1)
```
### Combined plot and table (folder)
The following for loop runs plotMetadata on each spectrum in the data_test list, which contains all the spectra in the folder. It will plot each spectrum and print its metadata below.
```{r folder-spectrum-metadata, message=FALSE, warning=FALSE, fig.width=8, fig.height=5, results='asis'}
for (i in seq_along(data_test)) {
  cat("#### Spectrum", i, ":", names(data_test)[i], "\n\n")
  data_i <- data_test[[i]]
  ab_data <- data_i$ab
  valid <- !is.null(ab_data) &&
    !is.null(ab_data$wavenumbers) && !is.null(ab_data$data) &&
    is.numeric(ab_data$wavenumbers) && is.numeric(ab_data$data) &&
    length(ab_data$wavenumbers) == length(ab_data$data) &&
    all(is.finite(ab_data$wavenumbers)) && all(is.finite(ab_data$data))
  if (valid) {
    plotSpectrum(data_i)
    print(metadataTable(data_i))
  } else {
    cat("Data not valid or missing for this spectrum.\n")
    str(ab_data)
  }
}
```
## Stacking multiple spectra
### Stacking re-runs
This code looks for files with the same sample number and stacks their spectra, to check whether reruns are consistent.
```{r stack-re-runs-new, include= TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=7}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  results = "markup",
  include = TRUE
)
library(opusreader2)
library(RColorBrewer)



data_test <- read_opus(dsn = corr_spectra)

# Helper to extract fields from filenames with the pattern
# AKS_suberin_[plant]_[washing-treatment]_[yymmdd]-[run].ext
extract_filename_fields <- function(filename) {
  parts <- strsplit(basename(filename), "_")[[1]]
  if (length(parts) >= 4) {
    list(
      plant = tolower(parts[3]),
      washing_treatment = tolower(parts[4])
    )
  } else {
    list(plant = NA, washing_treatment = NA)
  }
}

# Change this vector to c("plant") or c("washing_treatment") if you want
# to group on just one field instead of the plant-treatment pair.
grouping_fields <- c("plant", "washing_treatment")

file_fields <- lapply(names(data_test), extract_filename_fields)
plants <- vapply(file_fields, `[[`, character(1), "plant")
washing_treatments <- vapply(file_fields, `[[`, character(1), "washing_treatment")
group_df <- data.frame(
  plant = plants,
  washing_treatment = washing_treatments,
  stringsAsFactors = FALSE
)

group_keys <- do.call(paste, c(as.list(group_df[grouping_fields]), list(sep = "_")))
unique_combinations <- unique(group_keys[complete.cases(
  group_df[grouping_fields]
)])

# Plot stacked spectra for each unique combination
library(RColorBrewer)
for (combination in unique_combinations) {
  idx <- which(group_keys == combination)
  if (length(idx) > 1) {
    # Extract the configured grouping fields from the combination key
    parts <- strsplit(combination, "_")[[1]]
    plant <- if ("plant" %in% grouping_fields) parts[match("plant", grouping_fields)] else NA
    washing_treatment <- if ("washing_treatment" %in% grouping_fields) parts[match("washing_treatment", grouping_fields)] else NA

    spectra <- data_test[idx]
    colors <- brewer.pal(min(length(spectra), 8), "Set1")
    plot(NULL, xlim = rev(range(spectra[[1]]$ab$wavenumbers)),
         ylim = range(sapply(spectra, function(x) x$ab$data), na.rm = TRUE),
         xlab = "Wavenumber (cm⁻¹)", ylab = "Absorbance",
         main = paste("Stacked spectra for", paste(parts, collapse = " - ")),
         bty = "l")
    for (i in seq_along(spectra)) {
      ab_data <- spectra[[i]]$ab
      if (!is.null(ab_data) && !is.null(ab_data$wavenumbers) &&
            !is.null(ab_data$data)) {
        lines(ab_data$wavenumbers, ab_data$data, col = colors[i], lwd = 2)
      }
    }
    legend("topright",
           inset = c(-0.05, -0.02),
           legend = names(spectra),
           col = colors[seq_along(spectra)],
           lwd = 2,
           xpd = TRUE,
           bty = "n")
  }
}
```

### Stacking by plant \& washing-treatment
This code stacks spectra of the same plant and washing-treatment.
```{r stack-plant-wash, message=FALSE, warning=FALSE, fig.width=10, fig.height=7}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  results = "markup",
  include = TRUE
)
# Helper functions to extract plant and washing-treatment from filename
extract_filename_fields <- function(filename) {
  parts <- strsplit(basename(filename), "_")[[1]]
  if (length(parts) >= 4) {
    list(
      plant = tolower(parts[3]),
      washing_treatment = tolower(parts[4])
    )
  } else {
    list(plant = NA, washing_treatment = NA)
  }
}

file_fields <- lapply(names(data_test), extract_filename_fields)
plants <- vapply(file_fields, `[[`, character(1), "plant")
washing_treatments <- vapply(file_fields, `[[`, character(1), "washing_treatment")
group_df <- data.frame(
  plant = plants,
  washing_treatment = washing_treatments,
  stringsAsFactors = FALSE
)

library(viridisLite)
for (grouping_field in c("plant", "washing_treatment")) {
  # Unique grouping combinations
  group_keys <- group_df[[grouping_field]]
  combos <- unique(group_keys[!is.na(group_keys)])

  for (combo in combos) {
    idx <- which(group_keys == combo)
    if (length(idx) > 1) {
      spectra <- data_test[idx]
      colors <- magma(length(spectra), end = 0.9)
      plot(NULL, xlim = rev(range(spectra[[1]]$ab$wavenumbers)),
           ylim = range(sapply(spectra, function(x) x$ab$data), na.rm = TRUE),
           xlab = "Wavenumber (cm⁻¹)", ylab = "Absorbance",
           main = paste("Stacked spectra for", grouping_field, ":", combo),
           bty = "l")
      for (i in seq_along(spectra)) {
        ab_data <- spectra[[i]]$ab
        if (!is.null(ab_data) &&
              !is.null(ab_data$wavenumbers)
            && !is.null(ab_data$data)) {
          lines(ab_data$wavenumbers, ab_data$data, col = colors[i], lwd = 2)
        }
      }
      legend("topright",
             inset = c(-0.05, -0.02),
             legend = names(spectra),
             col = colors[seq_along(spectra)],
             lwd = 2,
             xpd = TRUE,
             bty = "n")
    }
  }
}
```